---
title: "Building Data Code"
author: "Mahal Bilaoen and Nancy Carmona"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
#-----setup options-----

# clear the workspace
rm(list = ls())

# knitr options, shows the code in the knitted HTML file 
knitr::opts_chunk$set(echo = TRUE)

```


```{r load_libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# load pacman, installing if needed
if (!require("pacman")) {install.packages("pacman")}

# load other packages, installing as needed
pacman::p_load(tidyverse, lubridate, stringr, tigris, sf, leaflet)


```

# Building Data Analysis 

All of the data sets associated with this file are all publicly available via SF Data Portal, which is the city of San Francisco's go to site to obtain data on the city. There is a lot of data to sift through and organize, especially if we're looking at different building types and where they are. The originial goal of downloading all of these data sets was to: 

1. Determine the A/C stock in the city 
2. Organize the city by building type
3. Identify where the city has the most heat exposure and determine vulnerability by A/C stock in a particular area and/or by building type/age 

# Loading in the data 

Building permit data is separated by before and after 01/01/2013, and it will be used to look at where any A/C related renovations are located. Tax roll and building audit data was mainly used to separate buildings by type, depending on how specific buildings were to be organized. Both data sets provide information about the buildings that could be useful. Due to data privacy, some of this data is available only on the zipcode level. 

```{r load_data}
#building permit data
before_2013 <- read.csv("Data/Building_Permits_before_January_1,_2013_20260626.csv")

after_2013 <- read.csv("Data/Building_Permits_filed_on_or_after_January_1,_2013_20260319.csv")

#tax data (building type)
tax_rolls <- read.csv("Data/Assessor_Historical_Secured_Property_Tax_Rolls_20260319.csv")

#energy compliance data (building type)
building_audits <- read.csv(
  "Data/Existing_Buildings_-_Basic_Info_and_Audit_Compliance_Status_20260319.csv"
)

```

 
The permit data comes in two files: permits before 2013 and permits after 2013. We first cleaned them separately so they have the same column names, then combine them.

```{r before_2013}

# create new data set 
# rename variables and select only the ones needed 

before_2013_clean <- before_2013 %>%
  mutate(
    data_source = "before_2013",
    
    block_clean = str_pad(as.character(Block), width = 4, side = "left", pad = "0"),
    lot_clean = str_pad(as.character(Lot), width = 3, side = "left", pad = "0"),
    parcel_id = str_c(block_clean, lot_clean),
    
    permit_number = as.character(Permit.Number),
    permit_type = as.character(Permit.Type),
    permit_type_definition = Permit.Type.Definition,
    
    permit_creation_date = ymd_hms(Permit.Creation.Date, quiet = TRUE),
    filed_date = ymd_hms(Filed.Date, quiet = TRUE),
    issued_date = ymd_hms(Issued.Date, quiet = TRUE),
    completed_date = ymd_hms(Completed.Date, quiet = TRUE),
    
    current_status = Current.Status,
    description = Description,
    
    block = as.character(Block),
    lot = as.character(Lot),
    
    street_number = as.character(Street.Number),
    street_name = Street.Name,
    street_suffix = Street.Suffix,
    
    zipcode = as.character(Zipcode),
    supervisor_district = as.character(supervisor_district),
    neighborhood = neighborhoods_analysis_boundaries,
    
    existing_use = Existing.Use,
    proposed_use = Proposed.Use,
    existing_units = as.numeric(Existing.Units),
    proposed_units = as.numeric(Proposed.Units),
    
    estimated_cost = as.numeric(Estimated.Cost),
    revised_cost = as.numeric(Revised.Cost),
    
    record_id = as.character(Record.ID),
    
    location = NA_character_,
    longitude = NA_real_,
    latitude = NA_real_
  ) %>%
  select(
    data_source,
    permit_number,
    permit_type,
    permit_type_definition,
    permit_creation_date,
    filed_date,
    issued_date,
    completed_date,
    current_status,
    description,
    block,
    lot,
    block_clean,
    lot_clean,
    parcel_id,
    street_number,
    street_name,
    street_suffix,
    zipcode,
    supervisor_district,
    neighborhood,
    existing_use,
    proposed_use,
    existing_units,
    proposed_units,
    estimated_cost,
    revised_cost,
    record_id,
    location,
    longitude,
    latitude
  )

```


```{r after_2013}

# create new data set 
# rename variables and select only the ones needed 

after_2013_clean <- after_2013 %>%
  mutate(
    data_source = "after_2013",
    
    block_clean = str_pad(as.character(Block), width = 4, side = "left", pad = "0"),
    lot_clean = str_pad(as.character(Lot), width = 3, side = "left", pad = "0"),
    parcel_id = str_c(block_clean, lot_clean),
    
    permit_number = as.character(Permit.Number),
    permit_type = as.character(Permit.Type),
    permit_type_definition = Permit.Type.Definition,
    
    permit_creation_date = ymd_hms(Permit.Creation.Date, quiet = TRUE),
    filed_date = ymd_hms(Filed.Date, quiet = TRUE),
    issued_date = ymd_hms(Issued.Date, quiet = TRUE),
    completed_date = ymd_hms(Completed.Date, quiet = TRUE),
    
    current_status = Current.Status,
    description = Description,
    
    block = as.character(Block),
    lot = as.character(Lot),
    
    street_number = as.character(Street.Number),
    street_name = Street.Name,
    street_suffix = Street.Suffix,
    
    zipcode = as.character(Zipcode),
    supervisor_district = as.character(supervisor_district),
    neighborhood = neighborhoods_analysis_boundaries,
    
    existing_use = Existing.Use,
    proposed_use = Proposed.Use,
    existing_units = as.numeric(Existing.Units),
    proposed_units = as.numeric(Proposed.Units),
    
    estimated_cost = as.numeric(Estimated.Cost),
    revised_cost = as.numeric(Revised.Cost),
    
    record_id = as.character(Record.ID),
    
    location = Location,
    longitude = as.numeric(str_extract(Location, "-\\d+\\.\\d+")),
    latitude = as.numeric(str_extract(Location, "\\d+\\.\\d+\\)") %>% str_remove("\\)"))
  ) %>%
  select(
    data_source,
    permit_number,
    permit_type,
    permit_type_definition,
    permit_creation_date,
    filed_date,
    issued_date,
    completed_date,
    current_status,
    description,
    block,
    lot,
    block_clean,
    lot_clean,
    parcel_id,
    street_number,
    street_name,
    street_suffix,
    zipcode,
    supervisor_district,
    neighborhood,
    existing_use,
    proposed_use,
    existing_units,
    proposed_units,
    estimated_cost,
    revised_cost,
    record_id,
    location,
    longitude,
    latitude
  )


```


```{r combine_permits}

# combine data sets 
building_permits_all <- bind_rows(
  before_2013_clean,
  after_2013_clean
)

```


# Identifying AC Updgrades 

On regex: regex (regular expression) is a function that looks at character strings and filters them out based on how exact you want the match to be. 

```{r cooling} 

# List of air cooling / HVAC-related terms
cooling_air_terms <- "(?i)\\b(air conditioning|A/C|AC|HVAC|heat pump|mini split|ductless|cooling|condenser|compressor|mechanical ventilation)\\b"

building_permits_all <- building_permits_all %>%
  mutate(
    cooling_air_update = if_else(
      str_detect(description, cooling_air_terms),
      1,
      0
    )
  )

# look at how many cooling updates
building_permits_all %>%
  count(cooling_air_update)

# view some examples 

building_permits_all %>%
  filter(cooling_air_update == 1) %>%
  select(
    data_source,
    permit_number,
    filed_date,
    description,
    cooling_air_update,
    zipcode,
    parcel_id
  ) %>%
  head(25)

```


## Tax Roll data -- lots of building definitions 

```{r tax_rolls}

# data cleaning for tax roll data

# Find the most recent tax roll year in the data
latest_tax_year <- tax_rolls %>%
  summarise(latest_year = max(Closed.Roll.Year, na.rm = TRUE)) %>%
  pull(latest_year)

latest_tax_year

# rename variables to simplify 
# only keep the variables needed for analysis since it is a really large file 

tax_rolls_clean <- tax_rolls %>%
  filter(Closed.Roll.Year == latest_tax_year) %>%
  mutate(
    parcel_id = as.character(Parcel.Number),
    
    tax_roll_year = Closed.Roll.Year,
    year_built = Year.Property.Built,
    property_use = Use.Definition,
    property_class = Property.Class.Code.Definition,
    number_of_units = parse_number(Number.of.Units),
    number_of_stories = Number.of.Stories,
    zoning_code = Zoning.Code,
    tax_roll_neighborhood = Analysis.Neighborhood,
    tax_roll_supervisor_district = as.character(Supervisor.District),
    parcel_location = the_geom
  ) %>%
  select(
    parcel_id,
    tax_roll_year,
    year_built,
    property_use,
    property_class,
    number_of_units,
    number_of_stories,
    zoning_code,
    tax_roll_neighborhood,
    tax_roll_supervisor_district,
    parcel_location
  ) %>%
  distinct(parcel_id, .keep_all = TRUE)


```
 
 
```{r merge_tax}

# merge the permit with tax data 
building_permits_tax <- building_permits_all %>%
  mutate(
    parcel_id = as.character(parcel_id)
  ) %>%
  left_join(tax_rolls_clean, by = "parcel_id")
```



## Building Type via Building Audits

Can use variable to examine how crowded a building may be. 


```{r audit_data}

# Clean building audit data
building_audits_clean <- building_audits %>%
  mutate(
    parcel_id = Parcel.Number,
    
    # Remove anything after a dash, such as "-B"
    parcel_id = str_remove(parcel_id, "-.*"),
    
    # Remove the slash between block and lot
    parcel_id = str_remove(parcel_id, "/"),
    
    # Rename useful audit variables
    audit_building_type = Property.Type...Self.Selected,
    audit_building_name = Building.Name,
    audit_building_address = Building.Address,
    audit_zipcode = as.character(Postal.Code),
    audit_floor_area = parse_number(Floor.Area),
    audit_year_built = Year.Built,
    audit_status = Energy.Audit.Status,
    audit_longitude = longitude,
    audit_latitude = latitude,
    audit_neighborhood = analysis_neighborhood,
    audit_supervisor_district = as.character(supervisor_district)
  ) %>%
  select(
    parcel_id,
    audit_building_type,
    audit_building_name,
    audit_building_address,
    audit_zipcode,
    audit_floor_area,
    audit_year_built,
    audit_status,
    audit_longitude,
    audit_latitude,
    audit_neighborhood,
    audit_supervisor_district
  ) %>%
  distinct(parcel_id, .keep_all = TRUE)

```

 

```{r combine_all}
# Combine permit + tax roll data with building audit data
final_building_data <- building_permits_tax %>%
  left_join(
    building_audits_clean,
    by = "parcel_id"
  )
```

##  Census Tracts

This section uses parcel coordinates from the tax roll data to assign each parcel to a census tract. We first match one row per parcel to a census tract, then join the tract information back to the full permit dataset.

Now we will add geographic information to the final building dataset!

We already have ZIP codes in the permit data. Some records also have ZIP codes from the building audit data. We will make one final ZIP code column using the permit ZIP code first. If that is missing, we will use the audit ZIP code.

For census tracts, we need latitude and longitude. Many older permit records do not have coordinates, so we will use the parcel location from the tax roll data instead. This gives us one location for each parcel.

```{r spatial_features}

# use tigris to get shapefile for san francisco 

options(tigris_use_cache = TRUE)

sf_tracts <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2020,
  cb = TRUE
)

# Check the result
sf_tracts %>%
  select(GEOID, NAME, geometry) %>%
  head()

```


```{r align_spatial}

# Add census tract information to the full final building dataset

final_building_data <- final_building_data %>%
  
  # Get longitude and latitude from parcel_location
  mutate(
    parcel_longitude = as.numeric(str_extract(parcel_location, "-\\d+\\.\\d+")),
    parcel_latitude = as.numeric(str_extract(parcel_location, "\\d+\\.\\d+\\)") %>%
                                   str_remove("\\)"))
  ) %>%
  
  # Keep one row per parcel for the spatial join
  select(
    parcel_id,
    parcel_longitude,
    parcel_latitude
  ) %>%
  filter(
    !is.na(parcel_longitude),
    !is.na(parcel_latitude)
  ) %>%
  distinct(parcel_id, .keep_all = TRUE) %>%
  
  # Convert parcel locations to spatial points
  st_as_sf(
    coords = c("parcel_longitude", "parcel_latitude"),
    crs = 4326,
    remove = FALSE
  ) %>%
  
  # Match each parcel point to a census tract
  st_join(
    st_transform(sf_tracts, 4326),
    join = st_intersects
  ) %>%
  
  # Keep only the tract information needed for the full dataset
  st_drop_geometry() %>%
  select(
    parcel_id,
    census_tract_geoid = GEOID,
    census_tract_name = NAME
  ) %>%
  
  # Join tract information back to the full permit/building dataset
  right_join(
    final_building_data,
    by = "parcel_id"
  )

# About 89% of permit records were matched to census tracts.
# Census tract assignment uses parcel locations from the tax roll data.
# Records without parcel locations could not be assigned a census tract.
```


 


```{r saving_datasets}

# Save cleaned intermediate datasets
write_csv(building_permits_all, "Data/building_permits_all.csv")
write_csv(tax_rolls_clean, "Data/tax_rolls_clean.csv")
write_csv(building_audits_clean, "Data/building_audits_clean.csv")

# Save merged datasets
write_csv(final_building_data, "Data/final_building_data.csv")

```

#### Questions for Team

Decisions to make: What is "new" housing? 
What year is the cut off?

```{r exploratory}
# Create analysis variables
final_building_data <- final_building_data %>%
  mutate(
    filed_year = year(filed_date),
    building_age = 2026 - year_built,
    
    newer_housing = if_else(
      year_built >= 2000, # what year are we considering new?
      1,
      0,
      missing = NA_real_
    )
  )

# Create one row per parcel

parcel_analysis_data <- final_building_data %>%
  group_by(parcel_id) %>%
  summarise(
    census_tract_geoid = first(census_tract_geoid),
    census_tract_name = first(census_tract_name),
    zipcode = first(zipcode),
    neighborhood = first(neighborhood),
    supervisor_district = first(supervisor_district),
    
    year_built = first(year_built),
    building_age = first(building_age),
    newer_housing = first(newer_housing),
    
    property_use = first(property_use),
    property_class = first(property_class),
    number_of_units = first(number_of_units),
    number_of_stories = first(number_of_stories),
    zoning_code = first(zoning_code),
    
    audit_building_type = first(audit_building_type),
    audit_floor_area = first(audit_floor_area),
    audit_status = first(audit_status),
    
    parcel_location = first(parcel_location),
    
    total_permits = n(),
    cooling_air_updates = sum(cooling_air_update, na.rm = TRUE),
    
    any_cooling_air_update = if_else(
      cooling_air_updates > 0,
      1,
      0
    ),
    
    .groups = "drop"
  )

# Overall cooling-air update summary

parcel_analysis_data %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    )
  )

# Cooling-air updates by neighborhood
cooling_by_neighborhood <- parcel_analysis_data %>%
  group_by(neighborhood) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_with_cooling_air_update))

cooling_by_neighborhood


```

#### Figure: Cooling by Neighborhood
```{r neighborhood_cooling}
ggplot(cooling_by_neighborhood, 
       aes(x = reorder(neighborhood, percent_with_cooling_air_update),
           y = percent_with_cooling_air_update)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Cooling-Air Updates by Neighborhood",
    x = "Neighborhood",
    y = "Percent of parcels with cooling-air update"
  )
```


```{r}

# Cooling-air updates by ZIP code
cooling_by_zipcode <- parcel_analysis_data %>%
  group_by(zipcode) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_with_cooling_air_update))

cooling_by_zipcode

# Cooling-air updates by census tract
cooling_by_census_tract <- parcel_analysis_data %>%
  group_by(census_tract_geoid, census_tract_name) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_with_cooling_air_update))

cooling_by_census_tract

# Older housing by neighborhood
housing_age_by_neighborhood <- parcel_analysis_data %>%
  group_by(neighborhood) %>%
  summarise(
    total_parcels = n(),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(median_building_age))

housing_age_by_neighborhood

# Neighborhoods with older housing and fewer cooling updates
cooling_age_by_neighborhood <- parcel_analysis_data %>%
  group_by(neighborhood) %>%
  summarise(
    total_parcels = n(),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(median_building_age), percent_with_cooling_air_update)

cooling_age_by_neighborhood

# Newer housing by neighborhood
newer_housing_by_neighborhood <- parcel_analysis_data %>%
  group_by(neighborhood) %>%
  summarise(
    total_parcels = n(),
    newer_housing_parcels = sum(newer_housing, na.rm = TRUE),
    percent_newer_housing = round(
      newer_housing_parcels / total_parcels * 100,
      2
    ),
    median_year_built = median(year_built, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_newer_housing))

newer_housing_by_neighborhood

# Cooling-air updates by property use
cooling_by_property_use <- parcel_analysis_data %>%
  group_by(property_use) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_with_cooling_air_update))

cooling_by_property_use

# Cooling-air updates by audit building type
# This only uses parcels that matched to the building audit dataset.

cooling_by_audit_building_type <- parcel_analysis_data %>%
  filter(!is.na(audit_building_type)) %>%
  group_by(audit_building_type) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_with_cooling_air_update))

cooling_by_audit_building_type



```



```{r}

# save data 

write_csv(parcel_analysis_data, "Data/parcel_analysis_data.csv")
write_csv(cooling_by_neighborhood, "Data/cooling_by_neighborhood.csv")
write_csv(cooling_by_zipcode, "Data/cooling_by_zipcode.csv")
write_csv(cooling_by_census_tract, "Data/cooling_by_census_tract.csv")
write_csv(housing_age_by_neighborhood, "Data/housing_age_by_neighborhood.csv")
write_csv(cooling_age_by_neighborhood, "Data/cooling_age_by_neighborhood.csv")
write_csv(newer_housing_by_neighborhood, "Data/newer_housing_by_neighborhood.csv")
write_csv(cooling_by_property_use, "Data/cooling_by_property_use.csv")
write_csv(cooling_by_audit_building_type, "Data/cooling_by_audit_building_type.csv")
```


# Exploratory Analysis


#### Summary Table by Parcel 

```{r table1}

# parcels are in the dataset
# parcels have cooling-air updates
# median year built, and median building age.

summary_table <- parcel_analysis_data %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE)
  )

summary_table

```

## Cooling improvements location

Where are cooling-related improvements occurring?

```{r cooling_neighborhood_summary}

# cooling-air updates by neighborhood

cooling_by_neighborhood <- parcel_analysis_data %>%
  group_by(neighborhood) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_with_cooling_air_update))

cooling_by_neighborhood


#  show only the top neighborhoods

cooling_by_neighborhood %>%
  slice_max(percent_with_cooling_air_update, n = 10)


```

Maybe need to remove GGP since people don't like there? 

```{r fig_neirborhoods}

# top neighborhoods for cooling-air updates
cooling_by_neighborhood %>%
  slice_max(percent_with_cooling_air_update, n = 15) %>%
  ggplot(aes(
    x = reorder(neighborhood, percent_with_cooling_air_update),
    y = percent_with_cooling_air_update
  )) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Neighborhoods for Cooling-Air Updates",
    x = "Neighborhood",
    y = "Percent of parcels with cooling-air update"
  )
```

Are older buildings less likely to have cooling-related updates?

```{r age_summary}

# building age vs. cooling-air update

parcel_analysis_data %>%
  filter(!is.na(building_age)) %>%
  mutate(
    cooling_air_update_label = if_else(
      any_cooling_air_update == 1,
      "Cooling-air update",
      "No cooling-air update"
    )
  ) %>%
  ggplot(aes(x = cooling_air_update_label, y = building_age)) +
  geom_boxplot() +
  labs(
    title = "Building Age by Cooling-Air Update Status",
    x = "",
    y = "Building age"
  )
```


 


### Building age association

Is building age associated with cooling-air updates?

Building age was significantly associated with cooling-air updates. The negative coefficient means that older buildings were slightly less likely to have a cooling-air update. This association was statistically significant, p < 0.001.

The effect size is very small, so building age alone probably does not explain much of the difference in cooling-air updates. Other factors, such as neighborhood, building type, property use, or historical disinvestment, may also be important.

```{r map_censustracts}

# using a simple linear model to estimate cooling update
# predictor is building age
# outcome is any cooling air update 

cooling_age_model <- glm(
  any_cooling_air_update ~ building_age,
  data = parcel_analysis_data,
  family = binomial
)

summary(cooling_age_model)

```


## Redlining Data

This section joins each parcel to a historical HOLC redlining grade using a spatial join.  Not all parcels will match because the redlining map does not cover every parcel in San Francisco.

```{r load_redlining, include = FALSE}

# save the file path in one object, to make it easier to edit later

redlining_file <- "~/Library/Mobile Documents/com~apple~CloudDocs/Documents/SFSU Faculty/PINC/PINC 2026/PSP-2026/Data/mappinginequality.json"

# read the spatial data 
redlining <- read_sf(redlining_file)

# check what kind of object this is
class(redlining)

# check the coordinate reference system
st_crs(redlining)

# check the geometry type
st_geometry_type(redlining)

# look at the first few rows of the data
head(redlining)

```

```{r check_redlining_data, include = FALSE}
 # select only San Francisco 
sf_redlining <- redlining %>% filter (city == "San Francisco") %>% 
  st_transform(4326)

# Note: Mapping requires longitude/latitude coordinates
# Our data are already WGS 84, but showing the transform step so you are aware


# view first rows 
head(sf_redlining)

# check the HOLC grades 
table(sf_redlining$grade, useNA = "ifany")

# remove NA
sf_redlining <- sf_redlining %>%
  filter(!is.na(grade))

# check the HOLC grades after filtering
table(sf_redlining$grade, useNA = "ifany")

```


```{r}
# Join parcel data to historical redlining grades

#  Add longitude and latitude from parcel_location
parcel_analysis_data <- parcel_analysis_data %>%
  mutate(
    parcel_longitude = as.numeric(str_extract(parcel_location, "-\\d+\\.\\d+")),
    parcel_latitude = as.numeric(
      str_extract(parcel_location, "\\d+\\.\\d+\\)") %>%
        str_remove("\\)")
    )
  )

#  Keep only parcels with coordinates for the spatial join
parcel_points_sf <- parcel_analysis_data %>%
  filter(
    !is.na(parcel_longitude),
    !is.na(parcel_latitude)
  ) %>%
  st_as_sf(
    coords = c("parcel_longitude", "parcel_latitude"),
    crs = 4326,
    remove = FALSE
  )

#  Make sure redlining data uses the same coordinate system
sf_redlining_4326 <- sf_redlining %>%
  st_transform(4326)

# Spatially join parcels to redlining polygons
sf_use_s2(FALSE)

# Clean redlining geometry
sf_redlining_4326_clean <- sf_redlining_4326 %>%
  st_make_valid()

# Join parcel points to redlining polygons
parcel_redlining_join <- parcel_points_sf %>%
  st_join(
    sf_redlining_4326_clean,
    join = st_intersects
  ) %>%
  st_drop_geometry() %>%
  select(
    parcel_id,
    redlining_grade = grade,
    redlining_category = category,
    redlining_area_id = area_id
  ) %>%
  distinct(parcel_id, .keep_all = TRUE)

# Turn s2 back on after the join
sf_use_s2(TRUE)

# Join redlining grades back to all parcels
parcel_analysis_redlining <- parcel_analysis_data %>%
  left_join(
    parcel_redlining_join,
    by = "parcel_id"
  )

```

Cooling-air updates by redlining grade:

```{r cooling_redlining}

# cooling updates by redlining grade 

cooling_by_redlining <- parcel_analysis_redlining %>%
  filter(!is.na(redlining_grade)) %>%
  group_by(redlining_grade, redlining_category) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(redlining_grade)

cooling_by_redlining

```
#### Mapping cooling-air updates by census tract
```{r}
# Summarize cooling-air updates by census tract
cooling_by_tract <- parcel_analysis_redlining %>%
  filter(!is.na(census_tract_geoid)) %>%
  group_by(census_tract_geoid) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  )

# Join tract summary to tract shapefile
cooling_tract_map <- sf_tracts %>%
  st_transform(4326) %>%
  left_join(
    cooling_by_tract,
    by = c("GEOID" = "census_tract_geoid")
  )

# Map percent of parcels with cooling-air updates
ggplot() +
  geom_sf(
    data = cooling_tract_map,
    aes(fill = percent_with_cooling_air_update),
    color = "white",
    linewidth = 0.1
  ) +
  labs(
    title = "Cooling-Air Updates by Census Tract",
    fill = "% of parcels"
  ) +
  theme_minimal()
```

#### Mapping Redlining

```{r}
 
# Make sure redlining data is in longitude/latitude for leaflet
sf_redlining_leaflet <- sf_redlining %>%
  st_transform(4326)

# Create HOLC color palette
holc_palette <- colorFactor(
  palette = c(
    "A" = "#76a865",  # Best
    "B" = "#7cb5bd",  # Still Desirable
    "C" = "#ffff00",  # Definitely Declining
    "D" = "#d9838d"   # Hazardous
  ),
  domain = sf_redlining_leaflet$grade
)

# Leaflet map of historical redlining grades
leaflet(sf_redlining_leaflet) %>%
  addTiles() %>%
  setView(
    lng = -122.4194,
    lat = 37.7749,
    zoom = 12
  ) %>%
  addPolygons(
    fillColor = ~holc_palette(grade),
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    popup = ~paste0(
      "<strong>HOLC Grade:</strong> ", grade, "<br>",
      "<strong>Category:</strong> ", category, "<br>",
      "<strong>Area ID:</strong> ", area_id
    )
  ) %>%
  addLegend(
    pal = holc_palette,
    values = ~grade,
    title = "HOLC Grade",
    opacity = 0.7
  )
 

cooling_by_redlining_area <- parcel_analysis_redlining %>%
  filter(!is.na(redlining_area_id)) %>%
  group_by(redlining_area_id, redlining_grade, redlining_category) %>%
  summarise(
    total_parcels = n(),
    parcels_with_cooling_air_update = sum(any_cooling_air_update, na.rm = TRUE),
    percent_with_cooling_air_update = round(
      parcels_with_cooling_air_update / total_parcels * 100,
      2
    ),
    .groups = "drop"
  )

redlining_cooling_map <- sf_redlining_leaflet %>%
  left_join(
    cooling_by_redlining_area,
    by = c("area_id" = "redlining_area_id")
  )

# map percent cooling-air updates by redlining area:

cooling_palette <- colorNumeric(
  palette = "YlOrRd",
  domain = redlining_cooling_map$percent_with_cooling_air_update,
  na.color = "transparent"
)


leaflet(redlining_cooling_map) %>%
  addTiles() %>%
  setView(
    lng = -122.4194,
    lat = 37.7749,
    zoom = 12
  ) %>%
  addPolygons(
    fillColor = ~cooling_palette(percent_with_cooling_air_update),
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    popup = ~paste0(
      "<strong>HOLC Grade:</strong> ", grade, "<br>",
      "<strong>Category:</strong> ", category, "<br>",
      "<strong>Total parcels:</strong> ", total_parcels, "<br>",
      "<strong>Parcels with cooling-air update:</strong> ",
      parcels_with_cooling_air_update, "<br>",
      "<strong>Percent with cooling-air update:</strong> ",
      percent_with_cooling_air_update, "%"
    )
  ) %>%
  addLegend(
    pal = cooling_palette,
    values = ~percent_with_cooling_air_update,
    title = "% with cooling-air update",
    opacity = 0.7
  )
```



