---
title: "Heat, Redlining, Cooling Updates, and Greening Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
#-----setup options-----

# clear the workspace
rm(list = ls())

# knitr options, shows the code in the knitted HTML file 
knitr::opts_chunk$set(echo = TRUE)

```

```{r load_libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# load pacman, installing if needed
if (!require("pacman")) {install.packages("pacman")}

# load other packages, installing as needed
pacman::p_load(tidyverse, broom, gt, gtsummary, viridis, patchwork)



```



# Introduction

This analysis uses three cleaned datasets which can be found in the Box Data folders.

1. `tract_year_full.csv`: census tract-year dataset  
2. `zip_year_full.csv`: ZIP code-year dataset  
3. `neighborhood_year_full.csv`: neighborhood-year dataset  


This project examines how housing age, historical redlining, greenness, and cooling-related building updates relate to heatwave exposure in San Francisco census tracts.

Aim 1: Examine whether housing age and newer development patterns are associated with heatwave exposure, tree planting, and cooling-related permits.

Aim 2: Describe temporal trends in heatwave exposure and greening activity from 2013 to 2022.

Aim 3: Assess whether historical redlining grade is associated with heatwave exposure, tree planting, and cooling-related permits.

```{r load_data}
# Read the cleaned analysis datasets from the Data folder.
# These files were created in the previous data cleaning script.

tract_year_full <- read_csv("Data/tract_year_full.csv")
zip_year_full <- read_csv("Data/zip_year_full.csv")
neighborhood_year_full <- read_csv("Data/neighborhood_year_full.csv")
```

```{r check_loaded_data}
# Check that the files loaded correctly.
# glimpse() is useful because it shows column names, data types, and example values.

glimpse(tract_year_full)
glimpse(zip_year_full)
glimpse(neighborhood_year_full)
```

Some data wrangling was done to change missing values are changed to 0. For example, if a tract-year has no recorded tree planting, we treat that as 0 trees planted.

```{r prepare_analysis_data}

# Clean census tract-year dataset for analysis.
# Remove 2012 because heat exposure data are not available for that year.
# Replace missing event counts with 0.
# Keep heat values as NA if any are still missing after removing 2012.

tract_year_analysis <- tract_year_full %>%
  filter(year != 2012) %>%
  mutate(
    total_permits = replace_na(total_permits, 0),
    cooling_air_updates = replace_na(cooling_air_updates, 0),
    trees_planted = replace_na(trees_planted, 0),
    any_cooling_air_update = cooling_air_updates > 0
  )

# Clean ZIP code-year dataset for analysis.

zip_year_analysis <- zip_year_full %>%
  filter(year != 2012) %>%
  mutate(
    total_permits = replace_na(total_permits, 0),
    cooling_air_updates = replace_na(cooling_air_updates, 0),
    trees_planted = replace_na(trees_planted, 0),
    any_cooling_air_update = cooling_air_updates > 0
  )

# Clean neighborhood-year dataset for analysis.

neighborhood_year_analysis <- neighborhood_year_full %>%
  filter(year != 2012) %>%
  mutate(
    total_permits = replace_na(total_permits, 0),
    cooling_air_updates = replace_na(cooling_air_updates, 0),
    trees_planted = replace_na(trees_planted, 0),
    any_cooling_air_update = cooling_air_updates > 0
  )
 

# check data
glimpse(tract_year_analysis)
glimpse(zip_year_analysis)
glimpse(neighborhood_year_analysis)


```

# Exploratory Data Checks

These checks help us understand the size and structure of the datasets before doing analysis.

```{r check_dataset_size}
# Check number of rows, geographies, and years.

tract_year_analysis %>%
  summarise(
    dataset = "Census tract-year",
    rows = n(),
    geographies = n_distinct(census_tract_geoid),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

zip_year_analysis %>%
  summarise(
    dataset = "ZIP code-year",
    rows = n(),
    geographies = n_distinct(zipcode),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

neighborhood_year_analysis %>%
  summarise(
    dataset = "Neighborhood-year",
    rows = n(),
    geographies = n_distinct(neighborhood),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )
```

```{r check_missingness}
# Check missingness in key variables.
# This helps identify variables that may need caution in analysis.

tract_year_analysis %>%
  summarise(
    rows = n(),
    missing_svi = sum(is.na(svi_overall)),
    missing_redlining_grade = sum(is.na(dominant_redlining_grade)),
    missing_building_age = sum(is.na(median_building_age)),
    missing_year_built = sum(is.na(median_year_built))
  )
```


## Overall Summary of Data

If this were a "Table 1" what would you want to show? Think about what the table is supposed to describe! 
A Table 1 helps the reader understand the study sample and compare important characteristics across groups. In this analysis, we can create tables by redlining grade, SVI quartile, and heat exposure quartile. For each table, consider which variables should be included, whether each variable should be summarized using a mean/standard deviation or median/interquartile range, and how missing values should be handled.  


```{r overall_summary}
# Overall summary across all census tract-years.

overall_summary <- tract_year_analysis %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    total_permits = sum(total_permits, na.rm = TRUE),
    total_cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(total_cooling_air_updates / total_permits * 100, 2),
    average_heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    average_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    total_trees_planted = sum(trees_planted, na.rm = TRUE),
    average_svi = mean(svi_overall, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE)
  )

overall_summary

```

## Yearly Trends

```{r yearly_summary}
# Summarize key variables by year.

yearly_summary <- tract_year_analysis %>%
  group_by(year) %>%
  summarise(
    total_permits = sum(total_permits, na.rm = TRUE),
    cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(cooling_air_updates / total_permits * 100, 2),
    average_heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    average_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    total_trees_planted = sum(trees_planted, na.rm = TRUE),
    average_svi = mean(svi_overall, na.rm = TRUE),
    .groups = "drop"
  )

yearly_summary
```

```{r plot_cooling_over_time}
# Plot cooling-air updates over time.

ggplot(yearly_summary, aes(x = year, y = cooling_air_updates)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Cooling-Air Updates Over Time",
    x = "Year",
    y = "Number of cooling-air updates"
  ) +
  scale_x_continuous(
  breaks = 2012:2022,
  minor_breaks = NULL) + 
  theme_minimal()
```

```{r plot_heat_over_time}
# Plot average 2 day heatwave events over time.

ggplot(yearly_summary, aes(x = year, y = average_heatwave_events_2day)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Heatwave Events Over Time",
    x = "Year",
    y = "Average 2 day heatwave events"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()
```

```{r plot_trees_over_time}
# Plot trees planted over time.

ggplot(yearly_summary, aes(x = year, y = total_trees_planted)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Trees Planted Over Time",
    x = "Year",
    y = "Trees planted"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()
```
 

## Analysis by Aim 

### Aim 1: Housing age, redevelopment, and heat exposure

Question: Do newer/redeveloped housing areas differ from older housing areas in heatwave exposure, tree planting, and cooling-related permits?

Main variables:
- Housing age (in relation to 2022) / redevelopment: median_year_built
- Heat exposure: heatwave_events_2day
- Greening activity: trees_planted
- Cooling adaptation: cooling_air_updates, percent_cooling_air, any_cooling_air_update

We are using 2000 as the cutoff for "newer housing." You may change this for your analysis! 

```{r building_age_variable}

#  Create "building era" variable

tract_year_analysis <- tract_year_analysis %>%
  mutate(
    building_era_binary = case_when(
      median_year_built < 2000 ~ "Older housing",
      median_year_built >= 2000 ~ "Newer housing",
      TRUE ~ NA_character_
    ),
    building_era_binary = factor(
      building_era_binary,
      levels = c(
        "Older housing",
        "Newer housing"
      )
    )
  ) %>%
  drop_na()


```


```{r aim1_table}

# compare older housing areas with newer housing areas on heatwave exposure, tree planting, and cooling-related permits.

aim1_descriptive <- tract_year_analysis %>%
  group_by(building_era_binary) %>%
  summarise(
    census_tracts = n_distinct(census_tract_geoid),
    
    median_year_built = median(median_year_built, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),
    
    median_heatwave_events_2day = median(heatwave_events_2day, na.rm = TRUE),
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    
    median_trees_planted = median(trees_planted, na.rm = TRUE),
    mean_trees_planted = mean(trees_planted, na.rm = TRUE),
    
    median_cooling_air_updates = median(cooling_air_updates, na.rm = TRUE),
    mean_cooling_air_updates = mean(cooling_air_updates, na.rm = TRUE),
    
    median_percent_cooling_air = median(percent_cooling_air, na.rm = TRUE),
    mean_percent_cooling_air = mean(percent_cooling_air, na.rm = TRUE),
    
    percent_with_any_cooling_update =
      mean(any_cooling_air_update, na.rm = TRUE) * 100,
    
    median_svi_overall = median(svi_overall, na.rm = TRUE),
    
    .groups = "drop"
  )

aim1_descriptive


# if you want to make a table, easier to "pivot" summary stats 

table1_building_era <- aim1_descriptive %>%
  filter(!is.na(building_era_binary)) %>%
  pivot_longer(
    cols = -building_era_binary,
    names_to = "characteristic",
    values_to = "value"
  ) %>%
  mutate(
    characteristic = recode(
      characteristic,
      census_tracts = "Census tracts",
      median_year_built = "Median year built",
      median_building_age = "Median building age",
      median_heatwave_events_2day = "Median 2-day heatwave events",
      mean_heatwave_events_2day = "Mean 2-day heatwave events",
      median_trees_planted = "Median trees planted",
      mean_trees_planted = "Mean trees planted",
      median_cooling_air_updates = "Median cooling/air updates",
      mean_cooling_air_updates = "Mean cooling/air updates",
      median_percent_cooling_air = "Median percent cooling/air",
      mean_percent_cooling_air = "Mean percent cooling/air",
      percent_with_any_cooling_update = "% with any cooling update",
      median_svi_overall = "Median SVI overall"
    )
  ) %>%
  pivot_wider(
    names_from = building_era_binary,
    values_from = value
  ) %>%
  gt() %>%
  fmt_number(
    columns = c(`Older housing`, `Newer housing`),
    decimals = 2
  ) %>%
  tab_header(
    title = "Table 1. Descriptive characteristics by building era"
  ) %>%
  cols_label(
    characteristic = "Characteristic",
    `Older housing` = "Older housing",
    `Newer housing` = "Newer housing"
  )

table1_building_era
 
```


```{r aim1_plots}

#  Descriptive plots

# Boxplots show the median and spread of the data.
# Note: Need to remove all the NAs!

ggplot(
  tract_year_analysis,
  aes(x = building_era_binary, y = heatwave_events_2day)
) +
  geom_boxplot() +
  labs(
    title = "2-Day Heatwave Events by Building Era",
    x = "Building era",
    y = "2-day heatwave events"
  ) +
  theme_minimal()

ggplot(
  tract_year_analysis,
  aes(x = building_era_binary, y = trees_planted)
) +
  geom_boxplot() +
  labs(
    title = "Trees Planted by Building Era",
    x = "Building era",
    y = "Trees planted"
  ) +
  theme_minimal()

ggplot(
  tract_year_analysis,
  aes(x = building_era_binary, y = cooling_air_updates)
) +
  geom_boxplot() +
  labs(
    title = "Cooling/Air Updates by Building Era",
    x = "Building era",
    y = "Cooling/air updates"
  ) +
  theme_minimal()
```

This unadjusted model tested whether 2-day heatwave events differed between older housing areas and newer housing areas.

* Older housing areas were the reference group.
* Older housing areas had an average of about 7.04 2-day heatwave events.
* Newer housing areas had about 2.19 more 2-day heatwave events than older housing areas.
* This difference was statistically significant, p = 0.030.

However, the R-squared was very low, about 0.003, meaning building era explained less than 1% of the variation in heatwave events.

```{r aim1_unadjusted}
# Inferential statistics: unadjusted model
#  model tests whether heatwave exposure differs by building era.

# Outcome: heatwave_events_2day
# Predictor: building_era_binary

aim1_heat_model_unadjusted <- lm(
  heatwave_events_2day ~ building_era_binary,
  data = tract_year_analysis
)

summary(aim1_heat_model_unadjusted)

```

#### Linear Heatwaves 
 
This adjusted linear regression model examined whether building era was associated with 2-day heatwave events after accounting for trees planted, cooling/air updates, and SVI.

* Older housing areas were the reference group.
* Newer housing areas had about **1.73 more 2-day heatwave events** than older housing areas.
* This difference was **not statistically significant at the 0.05 level**, but it was marginal, **p = 0.084**.
* Trees planted was significantly associated with 2-day heatwave events.
* Each additional tree planted was associated with about **0.08 more 2-day heatwave events**.
* Cooling/air updates were not significantly associated with heatwave events.
* SVI was not significantly associated with heatwave events.
* The overall model was statistically significant, **p < 0.001**, but the R-squared was low, meaning the model explained only a small amount of variation in heatwave exposure.

Overall, newer housing areas had somewhat higher 2-day heatwave exposure after adjustment, but the evidence was not strong enough to conclude a statistically significant difference. Trees planted was the strongest predictor in this model, but this should be interpreted as an association, not causation, because tree planting may have been targeted toward hotter areas.


```{r aim1_adjusted}

# Inferential statistics: adjusted model
# model tests whether building era is associated with heatwave exposure after accounting for selected tract characteristics.

# Outcome: heatwave_events_2day
# Predictors: building_era_binary, trees_planted, cooling_air_updates, svi_overall
 
aim1_heat_model_adjusted <- lm(
  heatwave_events_2day ~
    building_era_binary +
    trees_planted +
    cooling_air_updates +
    svi_overall,
  data = tract_year_analysis
)

summary(aim1_heat_model_adjusted)

```

#### Linear Heatwaves with Continuous Building Age

This adjusted linear regression model examined whether **median building age** was associated with 2-day heatwave events after accounting for trees planted, cooling/air updates, and SVI.

* Median building age was not significantly associated with 2-day heatwave events.
* Each additional year of median building age was associated with about **0.006 fewer 2-day heatwave events**, but this difference was not statistically significant.
* Trees planted was significantly associated with 2-day heatwave events.
* Each additional tree planted was associated with about **0.08 more 2-day heatwave events**.
* Cooling/air updates were not significantly associated with heatwave events.
* SVI was not significantly associated with heatwave events.
* The overall model was statistically significant, **p < 0.001**, but the R-squared was low, meaning the model explained only a small amount of variation in heatwave exposure.

Overall, median building age was not meaningfully associated with 2-day heatwave events after adjustment. Trees planted was the strongest predictor in this model, but this should be interpreted as an association, not causation, because tree planting may have been targeted toward hotter areas.


```{r aim1_cont}

# Inferential statistics: adjusted model
# model tests whether building era is associated with heatwave exposure after accounting for selected tract characteristics.

# Outcome: heatwave_events_2day
# Predictors: median_building_age, trees_planted, cooling_air_updates, svi_overall
aim1_heat_model_age_continuous <- lm(
  heatwave_events_2day ~ median_building_age + trees_planted + cooling_air_updates + svi_overall,
  data = tract_year_analysis
)

summary(aim1_heat_model_age_continuous)

```



### Aim 2: Greening and heatwave trends over time

Question: How did tree planting and heatwave exposure change from 2013 to 2022?


```{r aim2_table}
# Descriptive statistics by year
# annual trends in heatwave exposure and tree planting.
# They do not test whether the trends are statistically significant.

aim2_yearly_summary <- tract_year_analysis %>%
  group_by(year) %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    
    total_trees_planted = sum(trees_planted, na.rm = TRUE),
    median_trees_planted = median(trees_planted, na.rm = TRUE),
    mean_trees_planted = mean(trees_planted, na.rm = TRUE),
    
    median_heatwave_events_2day = median(heatwave_events_2day, na.rm = TRUE),
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    
    .groups = "drop"
  )

aim2_yearly_summary

```

#### Tree Planting Trends

```{r aim2_plots}

# Descriptive plot: tree planting over time
# plot shows whether total tree planting increased, decreased or fluctuated across years.

ggplot(
  aim2_yearly_summary,
  aes(x = year, y = total_trees_planted)
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Tree Planting Over Time",
    x = "Year",
    y = "Total trees planted"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()


# plot shows whether average 2-day heatwave events changed over time.

ggplot(
  aim2_yearly_summary,
  aes(x = year, y = mean_heatwave_events_2day)
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average 2-Day Heatwave Events Over Time",
    x = "Year",
    y = "Average 2-day heatwave events"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()



```


####  Linear Trees 

This linear regression model examined whether tree planting changed over time from 2013 to 2022.

* Year was significantly associated with trees planted, p < 0.001.
* The coefficient for year was -0.40.
* This means that, on average, each additional year was associated with about 0.40 fewer trees planted per census tract-year.
* The negative coefficient suggests that tree planting decreased over time.
* The overall model was statistically significant, but the R-squared was low, meaning year explained only a small amount of variation in tree planting.

Overall, tree planting showed a statistically significant downward trend over time, but year alone explained only a small portion of differences in tree planting across tract-years.


```{r}

# Inferential statistics: trend in tree planting
 #  model tests whether tree planting changed over time.

# Outcome: trees_planted
# Predictor: year

# Interpretation:
# The year coefficient tells us the average change in trees planted for each 1-year increase.

aim2_tree_trend_model <- lm(
  trees_planted ~ year,
  data = tract_year_analysis
)

summary(aim2_tree_trend_model)


```

 

### Aim 3: Historical redlining, heatwaves, greening, and cooling adaptation

Question: Do historically redlined areas differ in heatwave exposure, tree planting, and cooling-related permits?

 Main variables:
  Historical redlining: dominant_redlining_grade
 Heat exposure: heatwave_events_2day
 Greening activity: trees_planted
 Cooling adaptation: cooling_air_updates, percent_cooling_air, any_cooling_air_update
  Other tract characteristics: svi_overall, median_building_age


```{r}

# Need to assign HOLC grade!

#  Aim 3 analysis dataset
 
# We remove tract-years missing redlining grade because this aim compares redlining grades A, B, C, and D.
# Grade A is the reference group in the regression models.

aim3_data <- tract_year_analysis %>%
  filter(!is.na(dominant_redlining_grade)) %>%
  mutate(
    dominant_redlining_grade = factor(
      dominant_redlining_grade,
      levels = c("A", "B", "C", "D")
    )
  )
```

```{r}

#  Descriptive statistics by redlining grade
 
#   summaries describe patterns across redlining grades.
 
aim3_descriptive <- aim3_data %>%
  group_by(dominant_redlining_grade) %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    
    median_heatwave_events_2day = median(heatwave_events_2day, na.rm = TRUE),
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    
    median_trees_planted = median(trees_planted, na.rm = TRUE),
    mean_trees_planted = mean(trees_planted, na.rm = TRUE),
    
    median_cooling_air_updates = median(cooling_air_updates, na.rm = TRUE),
    mean_cooling_air_updates = mean(cooling_air_updates, na.rm = TRUE),
    
    median_percent_cooling_air = median(percent_cooling_air, na.rm = TRUE),
    mean_percent_cooling_air = mean(percent_cooling_air, na.rm = TRUE),
    
    percent_with_any_cooling_update =
      mean(any_cooling_air_update, na.rm = TRUE) * 100,
    
    median_svi_overall = median(svi_overall, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),
    
    .groups = "drop"
  )

aim3_descriptive


```


```{r}
 
#  Descriptive plot: heatwave exposure by redlining grade

#  plot compares 2-day heatwave events across HOLC grades.

ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = heatwave_events_2day)
) +
  geom_boxplot() +
  labs(
    title = "2-Day Heatwave Events by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "2-day heatwave events"
  ) +
  theme_minimal()


```



```{r}
 
# Descriptive plot: tree planting by redlining grade
#   plot compares tree planting across HOLC grades.

ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = trees_planted)
) +
  geom_boxplot() +
  labs(
    title = "Trees Planted by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "Trees planted"
  ) +
  theme_minimal()



```



```{r}

 #  Descriptive plot: cooling/air updates by redlining grade
 
#  plot compares cooling-related permit activity across HOLC grades.

ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = cooling_air_updates)
) +
  geom_boxplot() +
  labs(
    title = "Cooling/Air Updates by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "Cooling/air updates"
  ) +
  theme_minimal()

```


 SVI varied significantly by redlining grade

```{r aim3_svi_redlining}

 #  Descriptive plot: SVI by redlining grade
 
#  plot compares SVI by HOLC grades 

aim3_svi_by_redlining <- aim3_data %>%
  group_by(dominant_redlining_grade) %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    median_svi_overall = median(svi_overall, na.rm = TRUE),
    mean_svi_overall = mean(svi_overall, na.rm = TRUE),
    .groups = "drop"
  )

aim3_svi_by_redlining

# plot SVI by redlining
ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = svi_overall)
) +
  geom_boxplot() +
  labs(
    title = "Social Vulnerability by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "SVI overall"
  ) +
  theme_minimal()


# ANOVA
oneway.test(
  svi_overall ~ dominant_redlining_grade,
  data = aim3_data
)

```



#### Linear Redlining

The linear regression model examined 2-day heatwave events by redlining grade, SVI, median building age, and trees planted.

* Grade A was the reference group.
* Compared with grade A tracts, grades B, C, and D had slightly more 2-day heatwave events on average, but these differences were not statistically significant.
* SVI was not significantly associated with 2-day heatwave events.
* Median building age was not significantly associated with 2-day heatwave events.
* Trees planted was significantly associated with more 2-day heatwave events.
* Each additional tree planted was associated with about 0.09 more 2-day heatwave events.
* The overall model was statistically significant, p < 0.001, but the R-squared was low, meaning the model explained only a small amount of variation in heatwave exposure.

Overall, redlining grade was not significantly associated with 2-day heatwave events after accounting for SVI, median building age, and trees planted. Trees planted was the strongest predictor in this model. 


```{r aim3_redlining}
 
#  Inferential statistics: redlining and heatwave exposure
 
# This model tests whether 2-day heatwave events differ by redlining grade after accounting for SVI, median building age, and trees planted.

# Outcome: heatwave_events_2day

# Predictors: dominant_redlining_grade, svi_overall, median_building_age, trees_planted
 
# Grade A is the reference group. Coefficients for grades B, C, and D represent the average difference in 2-day heatwave events compared with grade A, holding the other variables constant.

aim3_heat_model <- lm(
  heatwave_events_2day ~
    dominant_redlining_grade +
    svi_overall +
    median_building_age +
    trees_planted,
  data = aim3_data
)

summary(aim3_heat_model)
```


#### Linear Trees

This linear regression model examined tree planting by redlining grade, heatwave exposure, SVI, and median building age.

* Grade A was the reference group.
* Compared with grade A tracts, grades B, C, and D had significantly fewer trees planted on average.
* Grade B tracts had about 5.2 fewer trees planted than grade A tracts.
* Grade C tracts had about 4.7 fewer trees planted than grade A tracts.
* Grade D tracts had about 4.8 fewer trees planted than grade A tracts.
* More 2-day heatwave events were significantly associated with more trees planted.
* Higher SVI was also significantly associated with more trees planted.
* Median building age was not statistically significant.
* The overall model was statistically significant, p < 0.001, but the R-squared was low, meaning the model explained a small amount of variation in tree planting.

Overall, tree planting was significantly lower in historically lower-graded areas compared with grade A areas, even after accounting for heatwave exposure, SVI, and median building age.


```{r aim3_trees}
 
#  Inferential statistics: redlining and tree planting
 # This model tests whether tree planting differs by redlining grade
# after accounting for heatwave exposure, SVI, and median building age.
 
# Outcome: trees_planted

# Predictors: dominant_redlining_grade, heatwave_events_2day, svi_overall, median_building_age

aim3_tree_model <- lm(
  trees_planted ~
    dominant_redlining_grade +
    heatwave_events_2day +
    svi_overall +
    median_building_age,
  data = aim3_data
)

summary(aim3_tree_model)
```

#### Linear Cooling

This adjusted linear regression model examined whether **redlining grade** was associated with the number of cooling/air updates after accounting for 2-day heatwave events, SVI, and median building age.

* Grade A was the reference group.
* Compared with grade A tracts, grades B, C, and D had fewer cooling/air updates on average.
* Grade B tracts had about **0.58 fewer cooling/air updates** than grade A tracts, but this difference was not statistically significant.
* Grade C tracts had about **0.49 fewer cooling/air updates** than grade A tracts, but this difference was not statistically significant.
* Grade D tracts had about **0.38 fewer cooling/air updates** than grade A tracts, but this difference was not statistically significant.
* 2-day heatwave events were not significantly associated with cooling/air updates.
* Higher SVI was significantly associated with more cooling/air updates.
* Older median building age was significantly associated with more cooling/air updates.
* The overall model was statistically significant, **p = 0.004**, but the R-squared was low, meaning the model explained only a small amount of variation in cooling/air updates.

Overall, redlining grade was not significantly associated with the number of cooling/air updates after adjustment. SVI and median building age were the strongest predictors in this model, but the model explained very little variation overall.



```{r aim3_cooling_linear}
 #  inferential statistics: redlining and cooling permits
 # This model tests whether cooling/air updates differ by redlining grade after accounting for heatwave exposure, SVI, and median building age.
 
# Outcome: cooling_air_updates
 
# Predictors: dominant_redlining_grade, heatwave_events_2day, svi_overall, median_building_age

aim3_cooling_model <- lm(
  cooling_air_updates ~
    dominant_redlining_grade +
    heatwave_events_2day +
    svi_overall +
    median_building_age,
  data = aim3_data
)

summary(aim3_cooling_model)
```





#### Logistic Cooling Outcome

This logistic regression model examined whether **redlining grade** was associated with the odds of having **any cooling/air update**, after accounting for 2-day heatwave events, SVI, and median building age.

* Grade A was the reference group.
* Compared with grade A tracts, grades B, C, and D had significantly lower odds of having any cooling/air update.
* Grade B tracts had lower odds of any cooling/air update, **OR = exp(-0.82) = 0.44**, p = 0.005.
* Grade C tracts had lower odds of any cooling/air update, **OR = exp(-0.64) = 0.53**, p = 0.017.
* Grade D tracts had lower odds of any cooling/air update, **OR = exp(-0.74) = 0.48**, p = 0.009.
* More 2-day heatwave events were significantly associated with lower odds of any cooling/air update.
* Higher SVI was significantly associated with lower odds of any cooling/air update.
* Older median building age was significantly associated with slightly higher odds of any cooling/air update.

Overall, historically lower-graded areas had significantly lower odds of having any cooling/air update compared with grade A areas, even after accounting for heatwave exposure, social vulnerability, and building age. This suggests that cooling-related adaptation may be less common in historically lower-graded areas.



```{r aim3_cooling_odds}

# inferential statistics: any cooling update
 # This logistic regression tests whether the odds of having any cooling/air update differ by redlining grade.

# Outcome: any_cooling_air_update

# Predictors: dominant_redlining_grade, heatwave_events_2day, svi_overall, median_building_age
 

aim3_any_cooling_model <- glm(
  any_cooling_air_update ~
    dominant_redlining_grade +
    heatwave_events_2day +
    svi_overall +
    median_building_age,
  data = aim3_data,
  family = binomial
)

summary(aim3_any_cooling_model)


```


```{r plot_aim3}
 
# Plot results from the cooling-update logistic regression model
 
# Are tract-years more or less likely to have any residential cooling update based on redlining grade, heatwave exposure, SVI, and building age?

# Interpretation:
# OR = 1: no association
# OR > 1: higher odds of any cooling update
# OR < 1: lower odds of any cooling update
 
# Convert coefficients to odds ratios
cooling_or_data <- tidy(
  aim3_any_cooling_model,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_clean = case_when(
      term == "dominant_redlining_gradeB" ~ "Redlining grade B vs A",
      term == "dominant_redlining_gradeC" ~ "Redlining grade C vs A",
      term == "dominant_redlining_gradeD" ~ "Redlining grade D vs A",
      term == "heatwave_events_2day" ~ "2-day Heatwave Events",
      term == "svi_overall" ~ "Social Vulnerability Index",
      term == "median_building_age" ~ "Median building age",
      TRUE ~ term
    ),
    term_clean = fct_rev(term_clean)
  )

# View odds ratio table
cooling_or_data

# Plot odds ratios
cooling_or_figure <- ggplot(
  cooling_or_data,
  aes(x = estimate, y = term_clean)
) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.8
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2,
    linewidth = 0.9,
    color = "#0072B2"
  ) +
  geom_point(
    size = 3.5,
    color = "#0072B2"
  ) +
  scale_x_log10(
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("0.25", "0.5", "1.0", "2.0", "4.0")
  ) +
  labs(
    title = "Predictors of Residential Cooling Updates",
    subtitle = "Adjusted odds ratios for any cooling update; lines show 95% confidence intervals",
    x = "Adjusted odds ratio",
    y = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

cooling_or_figure

# Save for poster
ggsave(
  filename = "cooling_update_logistic_model_figure.png",
  plot = cooling_or_figure,
  width = 9,
  height = 5.5,
  dpi = 300
)
```


## Extra Models - Comparing heat, trees, and cooling by HOLC grade

In adjusted logistic regression models, census tracts in historically lower-graded areas generally had lower odds of any residential cooling update compared with Grade A areas.

```{r}

# Outcomes need to be transformed to binary in order to use a logistic model 
# High heatwave exposure greater than median value
# Any tree planting per census tract year
# Any residential cooling update

 
# logistic model dataset

aim3_logit_data <- tract_year_analysis %>%
  mutate(
    # Binary outcome 1: high heatwave exposure
    # Defined as above the median number of 2-day heatwave events
    high_heatwave_exposure = heatwave_events_2day >
      median(heatwave_events_2day, na.rm = TRUE),

    # Binary outcome 2: any tree planting
    any_tree_planting = trees_planted > 0,

    # Binary outcome 3: any cooling update
    # This variable already exists in your tract_year_analysis
    any_cooling_update = any_cooling_air_update,

    # Make Grade A the reference group
    dominant_redlining_grade = factor(
      dominant_redlining_grade,
      levels = c("A", "B", "C", "D")
    )
  ) %>%
  filter(
    !is.na(dominant_redlining_grade),
    !is.na(high_heatwave_exposure),
    !is.na(any_tree_planting),
    !is.na(any_cooling_update),
    !is.na(svi_overall),
    !is.na(median_building_age)
  )

 
 # Logistic model: high heatwave exposure
 
aim3_heat_logit <- glm(
  high_heatwave_exposure ~ dominant_redlining_grade + svi_overall,
  data = aim3_logit_data,
  family = binomial
)

aim3_heat_or <- tidy(
  aim3_heat_logit,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  mutate(
    outcome = "High heatwave exposure"
  )

aim3_heat_or


 # Logistic model: any tree planting
 
aim3_tree_logit <- glm(
  any_tree_planting ~ dominant_redlining_grade + svi_overall,
  data = aim3_logit_data,
  family = binomial
)

aim3_tree_or <- tidy(
  aim3_tree_logit,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  mutate(
    outcome = "Any tree planting"
  )

aim3_tree_or


 # Logistic model: any residential cooling update
 
aim3_cooling_logit <- glm(
  any_cooling_update ~ dominant_redlining_grade + svi_overall + median_building_age,
  data = aim3_logit_data,
  family = binomial
)

aim3_cooling_or <- tidy(
  aim3_cooling_logit,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  mutate(
    outcome = "Any cooling update"
  )

aim3_cooling_or


 
# Combine redlining OR results for figure
 
aim3_or_plot_data <- bind_rows(
  aim3_heat_or,
  aim3_tree_or,
  aim3_cooling_or
) %>%
  filter(str_detect(term, "dominant_redlining_grade")) %>%
  mutate(
    redlining_grade = str_remove(term, "dominant_redlining_grade"),
    redlining_grade = case_when(
      redlining_grade == "B" ~ "B vs A",
      redlining_grade == "C" ~ "C vs A",
      redlining_grade == "D" ~ "D vs A",
      TRUE ~ redlining_grade
    ),
    redlining_grade = factor(
      redlining_grade,
      levels = c("B vs A", "C vs A", "D vs A")
    ),
    outcome = factor(
      outcome,
      levels = c(
        "High heatwave exposure",
        "Any tree planting",
        "Any cooling update"
      )
    )
  )

aim3_or_plot_data


# Plot odds ratios from aim3_or_plot_data
 
aim3_or_figure <- ggplot(
  aim3_or_plot_data,
  aes(
    x = estimate,
    y = outcome,
    color = outcome
  )
) +
  # OR = 1 means no difference from Grade A
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.8
  ) +
  # 95% confidence intervals
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.18,
    linewidth = 0.9
  ) +
  # Odds ratio point estimate
  geom_point(size = 3.5) +
  # Separate panels for each redlining comparison
  facet_wrap(~ redlining_grade, nrow = 1) +
  # Log scale is standard for odds ratios
  scale_x_log10(
    breaks = c(0.25, 0.5, 1, 2),
    labels = c("0.25", "0.5", "1.0", "2.0")
  ) +
  scale_color_manual(
    values = c(
      "High heatwave exposure" = "#D55E00",
      "Any tree planting" = "#009E73",
      "Any cooling update" = "#0072B2"
    )
  ) +
  labs(
    title = "Historical Redlining and Heat Adaptation Outcomes",
    subtitle = "Adjusted odds ratios compared with Grade A areas; lines show 95% confidence intervals",
    x = "Adjusted odds ratio compared with Grade A",
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    plot.subtitle = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

aim3_or_figure

ggsave(
  filename = "aim3_redlining_odds_ratio_figure.png",
  plot = aim3_or_figure,
  width = 13,
  height = 5.5,
  dpi = 300
)
```


 

