---
title: "Heat, Redlining, Cooling Updates, and Greening Analysis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
#-----setup options-----

# clear the workspace
rm(list = ls())

# knitr options, shows the code in the knitted HTML file 
knitr::opts_chunk$set(echo = TRUE)

```

```{r load_libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# load pacman, installing if needed
if (!require("pacman")) {install.packages("pacman")}

# load other packages, installing as needed
pacman::p_load(tidyverse, broom, gt, gtsummary, viridis, patchwork, corrplot, Hmisc)



```

# Introduction

This analysis uses three cleaned datasets which can be found in the Box
Data folders.

1.  `tract_year_full.csv`: census tract-year dataset\
2.  `zip_year_full.csv`: ZIP code-year dataset\
3.  `neighborhood_year_full.csv`: neighborhood-year dataset

This project examines how housing age, historical redlining, greenness,
and cooling-related building updates relate to heatwave exposure in San
Francisco census tracts.

Aim 1: Examine whether housing age and newer development patterns are
associated with heatwave exposure, tree planting, and cooling-related
permits.

Aim 2: Describe temporal trends in heatwave exposure and greening
activity from 2013 to 2022.

Aim 3: Assess whether historical redlining grade is associated with
heatwave exposure, tree planting, and cooling-related permits.

```{r load_data}
# Read the cleaned analysis datasets from the Data folder.
# These files were created in the previous data cleaning script.

tract_year_full <- read_csv("Data/tract_year_full.csv")
zip_year_full <- read_csv("Data/zip_year_full.csv")
neighborhood_year_full <- read_csv("Data/neighborhood_year_full.csv")
```

```{r check_loaded_data}
# Check that the files loaded correctly.
# glimpse() is useful because it shows column names, data types, and example values.

glimpse(tract_year_full)
glimpse(zip_year_full)
glimpse(neighborhood_year_full)
```

Some data wrangling was done to change missing values are changed to 0.
For example, if a tract-year has no recorded tree planting, we treat
that as 0 trees planted.

```{r prepare_analysis_data}

# Clean census tract-year dataset for analysis.
# Remove 2012 because heat exposure data are not available for that year.
# Replace missing event counts with 0.
# Keep heat values as NA if any are still missing after removing 2012.

tract_year_analysis <- tract_year_full %>%
  filter(year != 2012) %>%
  mutate(
    total_permits = replace_na(total_permits, 0),
    cooling_air_updates = replace_na(cooling_air_updates, 0),
    trees_planted = replace_na(trees_planted, 0),
    any_cooling_air_update = cooling_air_updates > 0
  )

# Clean ZIP code-year dataset for analysis.

zip_year_analysis <- zip_year_full %>%
  filter(year != 2012) %>%
  mutate(
    total_permits = replace_na(total_permits, 0),
    cooling_air_updates = replace_na(cooling_air_updates, 0),
    trees_planted = replace_na(trees_planted, 0),
    any_cooling_air_update = cooling_air_updates > 0
  )

# Clean neighborhood-year dataset for analysis.

neighborhood_year_analysis <- neighborhood_year_full %>%
  filter(year != 2012) %>%
  mutate(
    total_permits = replace_na(total_permits, 0),
    cooling_air_updates = replace_na(cooling_air_updates, 0),
    trees_planted = replace_na(trees_planted, 0),
    any_cooling_air_update = cooling_air_updates > 0
  )
 

# check data
glimpse(tract_year_analysis)
glimpse(zip_year_analysis)
glimpse(neighborhood_year_analysis)


```

# Exploratory Data Checks

These checks help us understand the size and structure of the datasets
before doing analysis.

```{r check_dataset_size}
# Check number of rows, geographies, and years.

tract_year_analysis %>%
  summarise(
    dataset = "Census tract-year",
    rows = n(),
    geographies = n_distinct(census_tract_geoid),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

zip_year_analysis %>%
  summarise(
    dataset = "ZIP code-year",
    rows = n(),
    geographies = n_distinct(zipcode),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

neighborhood_year_analysis %>%
  summarise(
    dataset = "Neighborhood-year",
    rows = n(),
    geographies = n_distinct(neighborhood),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )
```

```{r check_missingness}
# Check missingness in key variables.
# This helps identify variables that may need caution in analysis.

tract_year_analysis %>%
  summarise(
    rows = n(),
    missing_svi = sum(is.na(svi_overall)),
    missing_redlining_grade = sum(is.na(dominant_redlining_grade)),
    missing_building_age = sum(is.na(median_building_age)),
    missing_year_built = sum(is.na(median_year_built))
  )
```

## Overall Summary of Data

If this were a "Table 1" what would you want to show? Think about what
the table is supposed to describe! A Table 1 helps the reader understand
the study sample and compare important characteristics across groups. In
this analysis, we can create tables by redlining grade, SVI quartile,
and heat exposure quartile. For each table, consider which variables
should be included, whether each variable should be summarized using a
mean/standard deviation or median/interquartile range, and how missing
values should be handled.

```{r overall_summary}
# Overall summary across all census tract-years.

overall_summary <- tract_year_analysis %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    total_permits = sum(total_permits, na.rm = TRUE),
    total_cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(total_cooling_air_updates / total_permits * 100, 2),
    average_heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    average_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    total_trees_planted = sum(trees_planted, na.rm = TRUE),
    average_svi = mean(svi_overall, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE)
  )

overall_summary

```

## Yearly Trends

```{r yearly_summary}
# Summarize key variables by year.

yearly_summary <- tract_year_analysis %>%
  group_by(year) %>%
  summarise(
    total_permits = sum(total_permits, na.rm = TRUE),
    cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(cooling_air_updates / total_permits * 100, 2),
    average_heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    average_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    total_trees_planted = sum(trees_planted, na.rm = TRUE),
    average_svi = mean(svi_overall, na.rm = TRUE),
    .groups = "drop"
  )

yearly_summary
```

```{r plot_cooling_over_time}
# Plot cooling-air updates over time.

ggplot(yearly_summary, aes(x = year, y = cooling_air_updates)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Cooling-Air Updates Over Time",
    x = "Year",
    y = "Number of cooling-air updates"
  ) +
  scale_x_continuous(
  breaks = 2012:2022,
  minor_breaks = NULL) + 
  theme_minimal()
```

```{r plot_heat_over_time}
# Plot average 2 day heatwave events over time.

ggplot(yearly_summary, aes(x = year, y = average_heatwave_events_2day)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Heatwave Events Over Time",
    x = "Year",
    y = "Average 2 day heatwave events"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()
```

```{r plot_trees_over_time}
# Plot trees planted over time.

ggplot(yearly_summary, aes(x = year, y = total_trees_planted)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Trees Planted Over Time",
    x = "Year",
    y = "Trees planted"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()
```

## Analysis by Aim

### Aim 1: Housing age, redevelopment, and heat exposure

Question: Do newer/redeveloped housing areas differ from older housing
areas in heatwave exposure, tree planting, and cooling-related permits?

Main variables: - Housing age (in relation to 2022) / redevelopment:
median_year_built - Heat exposure: heatwave_events_2day - Greening
activity: trees_planted - Cooling adaptation: cooling_air_updates,
percent_cooling_air, any_cooling_air_update

We are using 2000 as the cutoff for "newer housing." You may change this
for your analysis!

```{r building_age_variable}

#  Create "building era" variable

tract_year_analysis <- tract_year_analysis %>%
  mutate(
    building_era_binary = case_when(
      median_year_built < 2000 ~ "Older housing",
      median_year_built >= 2000 ~ "Newer housing",
      TRUE ~ NA_character_
    ),
    building_era_binary = factor(
      building_era_binary,
      levels = c(
        "Older housing",
        "Newer housing"
      )
    )
  ) %>%
  drop_na()


```

```{r aim1_table}

# compare older housing areas with newer housing areas on heatwave exposure, tree planting, and cooling-related permits.

aim1_descriptive <- tract_year_analysis %>%
  group_by(building_era_binary) %>%
  summarise(
    census_tracts = n_distinct(census_tract_geoid),
    
    median_year_built = median(median_year_built, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),
    
    median_heatwave_events_2day = median(heatwave_events_2day, na.rm = TRUE),
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    
    median_trees_planted = median(trees_planted, na.rm = TRUE),
    mean_trees_planted = mean(trees_planted, na.rm = TRUE),
    
    median_cooling_air_updates = median(cooling_air_updates, na.rm = TRUE),
    mean_cooling_air_updates = mean(cooling_air_updates, na.rm = TRUE),
    
    median_percent_cooling_air = median(percent_cooling_air, na.rm = TRUE),
    mean_percent_cooling_air = mean(percent_cooling_air, na.rm = TRUE),
    
    percent_with_any_cooling_update =
      mean(any_cooling_air_update, na.rm = TRUE) * 100,
    
    median_svi_overall = median(svi_overall, na.rm = TRUE),
    
    .groups = "drop"
  )

aim1_descriptive


# if you want to make a table, easier to "pivot" summary stats 

table1_building_era <- aim1_descriptive %>%
  filter(!is.na(building_era_binary)) %>%
  pivot_longer(
    cols = -building_era_binary,
    names_to = "characteristic",
    values_to = "value"
  ) %>%
  mutate(
    characteristic = recode(
      characteristic,
      census_tracts = "Census tracts",
      median_year_built = "Median year built",
      median_building_age = "Median building age",
      median_heatwave_events_2day = "Median 2-day heatwave events",
      mean_heatwave_events_2day = "Mean 2-day heatwave events",
      median_trees_planted = "Median trees planted",
      mean_trees_planted = "Mean trees planted",
      median_cooling_air_updates = "Median cooling/air updates",
      mean_cooling_air_updates = "Mean cooling/air updates",
      median_percent_cooling_air = "Median percent cooling/air",
      mean_percent_cooling_air = "Mean percent cooling/air",
      percent_with_any_cooling_update = "% with any cooling update",
      median_svi_overall = "Median SVI overall"
    )
  ) %>%
  pivot_wider(
    names_from = building_era_binary,
    values_from = value
  ) %>%
  gt() %>%
  fmt_number(
    columns = c(`Older housing`, `Newer housing`),
    decimals = 2
  ) %>%
  tab_header(
    title = "Table 1. Descriptive characteristics by building era"
  ) %>%
  cols_label(
    characteristic = "Characteristic",
    `Older housing` = "Older housing",
    `Newer housing` = "Newer housing"
  )

table1_building_era
 
```

```{r aim1_plots}

#  Descriptive plots

# Boxplots show the median and spread of the data.
# Note: Need to remove all the NAs!

ggplot(
  tract_year_analysis,
  aes(x = building_era_binary, y = heatwave_events_2day)
) +
  geom_boxplot() +
  labs(
    title = "2-Day Heatwave Events by Building Era",
    x = "Building era",
    y = "2-day heatwave events"
  ) +
  theme_minimal()

ggplot(
  tract_year_analysis,
  aes(x = building_era_binary, y = trees_planted)
) +
  geom_boxplot() +
  labs(
    title = "Trees Planted by Building Era",
    x = "Building era",
    y = "Trees planted"
  ) +
  theme_minimal()

ggplot(
  tract_year_analysis,
  aes(x = building_era_binary, y = cooling_air_updates)
) +
  geom_boxplot() +
  labs(
    title = "Cooling/Air Updates by Building Era",
    x = "Building era",
    y = "Cooling/air updates"
  ) +
  theme_minimal()
```

This unadjusted model tested whether 2-day heatwave events differed
between older housing areas and newer housing areas.

- Older housing areas were the reference group.
- Older housing areas had an average of about 7.04 2-day heatwave
  events.
- Newer housing areas had about 2.19 more 2-day heatwave events than
  older housing areas.
- This difference was statistically significant, p = 0.030.

However, the R-squared was very low, about 0.003, meaning building era
explained less than 1% of the variation in heatwave events.

```{r aim1_unadjusted}
# Inferential statistics: unadjusted model
#  model tests whether heatwave exposure differs by building era.

# Outcome: heatwave_events_2day
# Predictor: building_era_binary

aim1_heat_model_unadjusted <- lm(
  heatwave_events_2day ~ building_era_binary,
  data = tract_year_analysis
)

summary(aim1_heat_model_unadjusted)

```

#### Linear Heatwaves

This adjusted linear regression model examined whether building era was
associated with 2-day heatwave events after accounting for trees
planted, cooling/air updates, and SVI.

- Older housing areas were the reference group.
- Newer housing areas had about **1.73 more 2-day heatwave events** than
  older housing areas.
- This difference was **not statistically significant at the 0.05
  level**, but it was marginal, **p = 0.084**.
- Trees planted was significantly associated with 2-day heatwave events.
- Each additional tree planted was associated with about **0.08 more
  2-day heatwave events**.
- Cooling/air updates were not significantly associated with heatwave
  events.
- SVI was not significantly associated with heatwave events.
- The overall model was statistically significant, **p \< 0.001**, but
  the R-squared was low, meaning the model explained only a small amount
  of variation in heatwave exposure.

Overall, newer housing areas had somewhat higher 2-day heatwave exposure
after adjustment, but the evidence was not strong enough to conclude a
statistically significant difference. Trees planted was the strongest
predictor in this model, but this should be interpreted as an
association, not causation, because tree planting may have been targeted
toward hotter areas.

```{r aim1_adjusted}

# Inferential statistics: adjusted model
# model tests whether building era is associated with heatwave exposure after accounting for selected tract characteristics.

# Outcome: heatwave_events_2day
# Predictors: building_era_binary, trees_planted, cooling_air_updates, svi_overall
 
aim1_heat_model_adjusted <- lm(
  heatwave_events_2day ~
    building_era_binary +
    trees_planted +
    cooling_air_updates +
    svi_overall,
  data = tract_year_analysis
)

summary(aim1_heat_model_adjusted)

```

#### Linear Heatwaves with Continuous Building Age

This adjusted linear regression model examined whether **median building
age** was associated with 2-day heatwave events after accounting for
trees planted, cooling/air updates, and SVI.

- Median building age was not significantly associated with 2-day
  heatwave events.
- Each additional year of median building age was associated with about
  **0.006 fewer 2-day heatwave events**, but this difference was not
  statistically significant.
- Trees planted was significantly associated with 2-day heatwave events.
- Each additional tree planted was associated with about **0.08 more
  2-day heatwave events**.
- Cooling/air updates were not significantly associated with heatwave
  events.
- SVI was not significantly associated with heatwave events.
- The overall model was statistically significant, **p \< 0.001**, but
  the R-squared was low, meaning the model explained only a small amount
  of variation in heatwave exposure.

Overall, median building age was not meaningfully associated with 2-day
heatwave events after adjustment. Trees planted was the strongest
predictor in this model, but this should be interpreted as an
association, not causation, because tree planting may have been targeted
toward hotter areas.

```{r aim1_cont}

# Inferential statistics: adjusted model
# model tests whether building era is associated with heatwave exposure after accounting for selected tract characteristics.

# Outcome: heatwave_events_2day
# Predictors: median_building_age, trees_planted, cooling_air_updates, svi_overall
aim1_heat_model_age_continuous <- lm(
  heatwave_events_2day ~ median_building_age + trees_planted + cooling_air_updates + svi_overall,
  data = tract_year_analysis
)

summary(aim1_heat_model_age_continuous)

```

### Aim 2: Greening and heatwave trends over time

Question: How did tree planting and heatwave exposure change from 2013
to 2022?

```{r aim2_table}
# Descriptive statistics by year
# annual trends in heatwave exposure and tree planting.
# They do not test whether the trends are statistically significant.

aim2_yearly_summary <- tract_year_analysis %>%
  group_by(year) %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    
    total_trees_planted = sum(trees_planted, na.rm = TRUE),
    median_trees_planted = median(trees_planted, na.rm = TRUE),
    mean_trees_planted = mean(trees_planted, na.rm = TRUE),
    
    median_heatwave_events_2day = median(heatwave_events_2day, na.rm = TRUE),
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    
    .groups = "drop"
  )

aim2_yearly_summary

```

#### Tree Planting Trends

```{r aim2_plots}

# Descriptive plot: tree planting over time
# plot shows whether total tree planting increased, decreased or fluctuated across years.

ggplot(
  aim2_yearly_summary,
  aes(x = year, y = total_trees_planted)
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Tree Planting Over Time",
    x = "Year",
    y = "Total trees planted"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()


# plot shows whether average 2-day heatwave events changed over time.

ggplot(
  aim2_yearly_summary,
  aes(x = year, y = mean_heatwave_events_2day)
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average 2-Day Heatwave Events Over Time",
    x = "Year",
    y = "Average 2-day heatwave events"
  ) +
    scale_x_continuous(
  breaks = 2013:2022,
  minor_breaks = NULL) + 
  theme_minimal()



```

#### Linear Trees

This linear regression model examined whether tree planting changed over
time from 2013 to 2022.

- Year was significantly associated with trees planted, p \< 0.001.
- The coefficient for year was -0.40.
- This means that, on average, each additional year was associated with
  about 0.40 fewer trees planted per census tract-year.
- The negative coefficient suggests that tree planting decreased over
  time.
- The overall model was statistically significant, but the R-squared was
  low, meaning year explained only a small amount of variation in tree
  planting.

Overall, tree planting showed a statistically significant downward trend
over time, but year alone explained only a small portion of differences
in tree planting across tract-years.

```{r}

# Inferential statistics: trend in tree planting
 #  model tests whether tree planting changed over time.

# Outcome: trees_planted
# Predictor: year

# Interpretation:
# The year coefficient tells us the average change in trees planted for each 1-year increase.

aim2_tree_trend_model <- lm(
  trees_planted ~ year,
  data = tract_year_analysis
)

summary(aim2_tree_trend_model)


```

### Aim 3: Historical redlining, heatwaves, greening, and cooling adaptation

Question: Do historically redlined areas differ in heatwave exposure,
tree planting, and cooling-related permits?

Main variables: Historical redlining: dominant_redlining_grade Heat
exposure: heatwave_events_2day Greening activity: trees_planted Cooling
adaptation: cooling_air_updates, percent_cooling_air,
any_cooling_air_update Other tract characteristics: svi_overall,
median_building_age

```{r aim3_data}

# Need to assign HOLC grade!

#  Aim 3 analysis dataset
 
# We remove tract-years missing redlining grade because this aim compares redlining grades A, B, C, and D.
# Grade A is the reference group in the regression models.

aim3_data <- tract_year_analysis %>%
  filter(!is.na(dominant_redlining_grade)) %>%
  mutate(
    dominant_redlining_grade = factor(
      dominant_redlining_grade,
      levels = c("A", "B", "C", "D")
    )
  )
```

```{r aim3_descriptive}

#  Descriptive statistics by redlining grade
 
#   summaries describe patterns across redlining grades.
 
aim3_descriptive <- aim3_data %>%
  group_by(dominant_redlining_grade) %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    
    median_heatwave_events_2day = median(heatwave_events_2day, na.rm = TRUE),
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    
    median_trees_planted = median(trees_planted, na.rm = TRUE),
    mean_trees_planted = mean(trees_planted, na.rm = TRUE),
    
    median_cooling_air_updates = median(cooling_air_updates, na.rm = TRUE),
    mean_cooling_air_updates = mean(cooling_air_updates, na.rm = TRUE),
    
    median_percent_cooling_air = median(percent_cooling_air, na.rm = TRUE),
    mean_percent_cooling_air = mean(percent_cooling_air, na.rm = TRUE),
    
    percent_with_any_cooling_update =
      mean(any_cooling_air_update, na.rm = TRUE) * 100,
    
    median_svi_overall = median(svi_overall, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),
    
    .groups = "drop"
  )

aim3_descriptive


```

### Table 1

In these tables we used Kruskal-Wallis tests to compare median values
across SVI quartiles and HOLC/redlining grades because the variables
were not normally distributed.

In these tables, we used chi-square tests to compare categorical
variables across groups. For example, the chi-square test assessed
whether the distribution of HOLC/redlining grades differed across SVI
quartiles.

```{r aim3_table1}
 
# Prepare variables for Table 1
 table1_data <- aim3_data %>%
  mutate(
    # SVI quartiles: Q1 = lowest vulnerability, Q4 = highest vulnerability
    svi_quartile = ntile(svi_overall, 4),
    svi_quartile = factor(
      svi_quartile,
      levels = c(1, 2, 3, 4),
      labels = c(
        "Q1: Lowest SVI",
        "Q2",
        "Q3",
        "Q4: Highest SVI"
      )
    ),

    # Make HOLC/redlining grade easier to read
    dominant_redlining_grade = factor(
      dominant_redlining_grade,
      levels = c("A", "B", "C", "D"),
      labels = c(
        "A: Best",
        "B: Still desirable",
        "C: Definitely declining",
        "D: Hazardous"
      )
    )
  )
```

#### Table 1 - SVI

- Historical redlining grade varied significantly by SVI quartile The
  distribution of HOLC/redlining grades was significantly different
  across SVI quartiles, p \< 0.001.

- The highest SVI quartile had the largest percentage of historically
  redlined “D: Hazardous” areas In Q4, the highest SVI group, 60% of
  census tract-years were classified as D: Hazardous, compared with 18%
  in Q1, the lowest SVI group. No “A: Best” HOLC areas were in the
  highest SVI quartile In Q4, 0% were classified as A: Best, compared
  with 9.2% in Q1.

- Heatwave exposure did not differ significantly by SVI quartile\*
  Median 2-day heatwave events were similar across SVI quartiles, around
  7–8 events, and the difference was not statistically significant, p \>
  0.9.

- Cooling/air updates differed significantly by SVI quartile Cooling/air
  updates varied by SVI quartile, p \< 0.001. The median was 0 updates
  across all quartiles, but the lower SVI quartiles had slightly higher
  upper ranges.

- Tree planting did not significantly differ by SVI quartile Median
  trees planted were similar across groups, ranging from 3 to 4 trees,
  and the difference was not statistically significant, p = 0.083.

- Median building age differed significantly by SVI quartile Building
  age varied by SVI quartile, p \< 0.001. Lower SVI areas had older
  median building ages, while the highest SVI areas had younger median
  building ages.

```{r table1_SVI}

# make a table 1 by SVI quartile
table1_by_svi <- table1_data %>%
  select(
    svi_quartile,
    dominant_redlining_grade,
    heatwave_events_2day,
    cooling_air_updates,
    trees_planted,
    median_building_age
  ) %>%
  tbl_summary(
    by = svi_quartile,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no",
    label = list(
      dominant_redlining_grade ~ "HOLC/redlining grade",
      heatwave_events_2day ~ "2-day heatwave events",
      cooling_air_updates ~ "Cooling/air updates",
      trees_planted ~ "Trees planted",
      median_building_age ~ "Median building age"
    )
  ) %>%
  add_n() %>%
  add_p(
    test = list(
      all_continuous() ~ "kruskal.test",
      all_categorical() ~ "chisq.test"
    )
  ) %>%
  bold_labels()

table1_by_svi
```

#### Table 1 - HOLC Grade

- Social Vulnerability Index differed significantly by HOLC/redlining
  grade SVI was significantly different across HOLC grades, p \< 0.001.
  D-grade areas had the highest social vulnerability The median SVI was
  highest in D: Hazardous areas, with a median of 0.45, compared with
  0.10 in A: Best areas.

- Heatwave events did not differ significantly by HOLC/redlining grade
  Median 2-day heatwave events were similar across HOLC grades, ranging
  from 7 to 8 events, p = 0.5. Cooling/air updates differed
  significantly by HOLC/redlining grade

- Cooling/air updates varied across HOLC grades, p \< 0.001. However,
  the median number of updates was 0 in every HOLC group, meaning many
  census tract-years had no cooling/air updates.

- Trees planted differed significantly by HOLC/redlining grade Tree
  planting varied across HOLC grades, p \< 0.001. A-grade areas had the
  highest median number of trees planted A: Best areas had a median of 8
  trees planted, compared with 3–4 trees in B, C, and D areas. Median
  building age differed significantly by HOLC/redlining grade

- Building age varied across HOLC grades, p \< 0.001. D-grade areas had
  among the oldest median building ages D: Hazardous areas had a median
  building age of 105 years, compared with 99 years in A-grade areas and
  90 years in B-grade areas.

 



```{r}

# make a table 1 by Grade 

table1_by_holc <- table1_data %>%
  select(
    dominant_redlining_grade,
    svi_overall,
    heatwave_events_2day,
    cooling_air_updates,
    trees_planted,
    median_building_age
  ) %>%
  tbl_summary(
    by = dominant_redlining_grade,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})"
    ),
    missing = "no",
    label = list(
      svi_overall ~ "Social Vulnerability Index",
      heatwave_events_2day ~ "2-day heatwave events",
      cooling_air_updates ~ "Cooling/air updates",
      trees_planted ~ "Trees planted",
      median_building_age ~ "Median building age"
    )
  ) %>%
  add_n() %>%
  add_p(
    test = list(
      all_continuous() ~ "kruskal.test"
    )
  ) %>%
  bold_labels()

table1_by_holc
```

#### Correlations

Please include rho and p values for the results!

Using the package corrplot to visualize the correlation

<https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

We will use Spearman because the data is not normally distributed.

```{r}
# Run correlation test
building_age_cooling <- cor.test(aim3_data$median_building_age,
                                 aim3_data$cooling_air_updates,
                                 method = "spearman")
print(building_age_cooling)

building_age_cooling_percent <- cor.test(aim3_data$median_building_age,
                                         aim3_data$percent_cooling_air,
                                         method = "spearman")
print(building_age_cooling_percent)

```



```{r correlation_subset}

# select only the variables that will go into your plot 

correlation_data <- aim3_data %>%
  select(
    heatwave_events_2day,
    svi_overall,
    trees_planted,
    cooling_air_updates,
    median_building_age
  )

#  Create Spearman correlation matrix
correlation_matrix <- cor(
  correlation_data,
  use = "pairwise.complete.obs",
  method = "spearman"
)


# Plot correlation matrix
corrplot(
  correlation_matrix,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  number.cex = 0.7,
  tl.cex = 0.8,
  tl.col = "black",
  tl.srt = 45,
  diag = FALSE
)

```

- Trees planted and heatwave events were significantly positively
  correlated Spearman rho = 0.21, p \< 0.001. Census tracts with more
  2-day heatwave events tended to have more trees planted. The
  relationship was statistically significant but weak.

- Median building age and SVI were significantly negatively correlated
  Spearman rho = -0.41, p \< 0.001.Census tracts with higher social
  vulnerability tended to have lower median building age, or census
  tracts with older buildings tended to have lower SVI. This
  relationship was moderate in strength.

- Cooling/air updates and SVI were significantly negatively correlated
  Spearman rho = -0.17, p \< 0.001. Census tracts with higher social
  vulnerability tended to have fewer cooling or air-related building
  updates. The relationship was statistically significant but weak.

- Median building age and cooling/air updates were significantly
  positively correlated Spearman rho = 0.14, p \< 0.001. Census tracts
  with older buildings tended to have slightly more cooling or
  air-related updates. The relationship was statistically significant
  but weak.

- Cooling/air updates and heatwave events were significantly negatively
  correlated Spearman rho = -0.08, p \< 0.01. Census tracts with more
  heatwave events tended to have slightly fewer cooling or air-related
  updates. The relationship was statistically significant, but very
  weak.

- Median building age and trees planted were significantly positively
  correlated Spearman rho = 0.07, p \< 0.01. Census tracts with older
  buildings tended to have slightly more trees planted. The relationship
  was statistically significant, but very weak.

```{r correlation_pvalue}

# To view which correlations are significant create Spearman correlation matrix and p-value matrix

correlation_results <- Hmisc::rcorr(
  as.matrix(correlation_data),
  type = "spearman"
)

# Correlation coefficients
correlation_matrix <- correlation_results$r

# P-values for each correlation
p_value_matrix <- correlation_results$P

# print p-values to view values 
p_value_matrix

```

```{r}
 
#  Descriptive plot: heatwave exposure by redlining grade

#  plot compares 2-day heatwave events across HOLC grades.

ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = heatwave_events_2day)
) +
  geom_boxplot() +
  labs(
    title = "2-Day Heatwave Events by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "2-day heatwave events"
  ) +
  theme_minimal()


```

#### Box plot: Trees by Redlining
```{r}
 
# Descriptive plot: tree planting by redlining grade
#   plot compares tree planting across HOLC grades.

 

library(ggplot2)
library(scales)  # for alpha()

holc_colors <- c(
  "A" = alpha("green", 0.6),
  "B" = alpha("blue", 0.6),
  "C" = alpha("yellow", 0.6),
  "D" = alpha("red", 0.6)
)


ggplot(
  aim3_data,
  aes(
    x = dominant_redlining_grade,
    y = trees_planted,
    fill = dominant_redlining_grade
  )
) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = holc_colors) +
  labs(
    title = "Trees Planted by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "Trees planted",
    fill = "Redlining grade"
  ) +
  theme_minimal() + 
     theme(
    plot.title = element_text(hjust = 0.5)
  )


```

```{r}

 #  Descriptive plot: cooling/air updates by redlining grade
 
#  plot compares cooling-related permit activity across HOLC grades.

ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = cooling_air_updates)
) +
  geom_boxplot() +
  labs(
    title = "Cooling/Air Updates by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "Cooling/air updates"
  ) +
  theme_minimal()

```

SVI varied significantly by redlining grade

```{r aim3_svi_redlining}

 #  Descriptive plot: SVI by redlining grade
 
#  plot compares SVI by HOLC grades 

aim3_svi_by_redlining <- aim3_data %>%
  group_by(dominant_redlining_grade) %>%
  summarise(
    tract_years = n(),
    census_tracts = n_distinct(census_tract_geoid),
    median_svi_overall = median(svi_overall, na.rm = TRUE),
    mean_svi_overall = mean(svi_overall, na.rm = TRUE),
    .groups = "drop"
  )

aim3_svi_by_redlining

# plot SVI by redlining
ggplot(
  aim3_data,
  aes(x = dominant_redlining_grade, y = svi_overall)
) +
  geom_boxplot() +
  labs(
    title = "Social Vulnerability by Historical Redlining Grade",
    x = "Dominant redlining grade",
    y = "SVI overall"
  ) +
  theme_minimal()


# ANOVA
oneway.test(
  svi_overall ~ dominant_redlining_grade,
  data = aim3_data
)

```

#### Linear Redlining

The linear regression model examined 2-day heatwave events by redlining
grade, SVI, median building age, and trees planted.

- Grade A was the reference group.
- Compared with grade A tracts, grades B, C, and D had slightly more
  2-day heatwave events on average, but these differences were not
  statistically significant.
- SVI was not significantly associated with 2-day heatwave events.
- Median building age was not significantly associated with 2-day
  heatwave events.
- Trees planted was significantly associated with more 2-day heatwave
  events.
- Each additional tree planted was associated with about 0.09 more 2-day
  heatwave events.
- The overall model was statistically significant, p \< 0.001, but the
  R-squared was low, meaning the model explained only a small amount of
  variation in heatwave exposure.

Overall, redlining grade was not significantly associated with 2-day
heatwave events after accounting for SVI, median building age, and trees
planted. Trees planted was the strongest predictor in this model.

```{r aim3_redlining}
 
#  Inferential statistics: redlining and heatwave exposure
 
# This model tests whether 2-day heatwave events differ by redlining grade after accounting for SVI, median building age, and trees planted.

# Outcome: heatwave_events_2day

# Predictors: dominant_redlining_grade, svi_overall, median_building_age, trees_planted
 
# Grade A is the reference group. Coefficients for grades B, C, and D represent the average difference in 2-day heatwave events compared with grade A, holding the other variables constant.

aim3_heat_model <- lm(
  heatwave_events_2day ~
    dominant_redlining_grade +
    svi_overall +
    median_building_age +
    trees_planted,
  data = aim3_data
)

summary(aim3_heat_model)
```

#### Linear Trees

This linear regression model examined tree planting by redlining grade,
heatwave exposure, SVI, and median building age.

- Grade A was the reference group.
- Compared with grade A tracts, grades B, C, and D had significantly
  fewer trees planted on average.
- Grade B tracts had about 5.2 fewer trees planted than grade A tracts.
- Grade C tracts had about 4.7 fewer trees planted than grade A tracts.
- Grade D tracts had about 4.8 fewer trees planted than grade A tracts.
- More 2-day heatwave events were significantly associated with more
  trees planted.
- Higher SVI was also significantly associated with more trees planted.
- Median building age was not statistically significant.
- The overall model was statistically significant, p \< 0.001, but the
  R-squared was low, meaning the model explained a small amount of
  variation in tree planting.

Overall, tree planting was significantly lower in historically
lower-graded areas compared with grade A areas, even after accounting
for heatwave exposure, SVI, and median building age.

```{r aim3_trees}
 
#  Inferential statistics: redlining and tree planting
 # This model tests whether tree planting differs by redlining grade
# after accounting for heatwave exposure, SVI, and median building age.
 
# Outcome: trees_planted

# Predictors: dominant_redlining_grade, heatwave_events_2day, svi_overall, median_building_age

aim3_tree_model <- lm(
  trees_planted ~
    dominant_redlining_grade +
    heatwave_events_2day +
    svi_overall +
    median_building_age,
  data = aim3_data
)

summary(aim3_tree_model)
```

#### Linear Cooling

This adjusted linear regression model examined whether **redlining
grade** was associated with the number of cooling/air updates after
accounting for 2-day heatwave events, SVI, and median building age.

- Grade A was the reference group.
- Compared with grade A tracts, grades B, C, and D had fewer cooling/air
  updates on average.
- Grade B tracts had about **0.58 fewer cooling/air updates** than grade
  A tracts, but this difference was not statistically significant.
- Grade C tracts had about **0.49 fewer cooling/air updates** than grade
  A tracts, but this difference was not statistically significant.
- Grade D tracts had about **0.38 fewer cooling/air updates** than grade
  A tracts, but this difference was not statistically significant.
- 2-day heatwave events were not significantly associated with
  cooling/air updates.
- Higher SVI was significantly associated with more cooling/air updates.
- Older median building age was significantly associated with more
  cooling/air updates.
- The overall model was statistically significant, **p = 0.004**, but
  the R-squared was low, meaning the model explained only a small amount
  of variation in cooling/air updates.

Overall, redlining grade was not significantly associated with the
number of cooling/air updates after adjustment. SVI and median building
age were the strongest predictors in this model, but the model explained
very little variation overall.

```{r aim3_cooling_linear}
 #  inferential statistics: redlining and cooling permits
 # This model tests whether cooling/air updates differ by redlining grade after accounting for heatwave exposure, SVI, and median building age.
 
# Outcome: cooling_air_updates
 
# Predictors: dominant_redlining_grade, heatwave_events_2day, svi_overall, median_building_age

aim3_cooling_model <- lm(
  cooling_air_updates ~
    dominant_redlining_grade +
    heatwave_events_2day +
    svi_overall +
    median_building_age,
  data = aim3_data
)

summary(aim3_cooling_model)
```

#### Logistic Cooling Outcome

Note: We should check our data before building a model to determine what
model is appropriate.

```{r hist, echo = FAlSE, warning = FALSE}

# ---- we check the normality of the outcome to help determine what model to use ---- 

# if the histogram is skewed or the Q–Q plot does not match the 1:1 line, then the data is likely not normally distributed 

ggplot(aim3_data, aes(x = cooling_air_updates)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Histogram of tract-level cooling updates")

# we use sample as the argument because we are using "stat_qq"

ggplot(aim3_data, aes(sample = cooling_air_updates)) +
  stat_qq() + stat_qq_line() +
  labs(title = "Q–Q plot versus Normal")


```

This logistic regression model examined whether **redlining grade** was
associated with the odds of having **any cooling/air update**, after
accounting for 2-day heatwave events, SVI, and median building age.

- Grade A was the reference group.
- Compared with grade A tracts, grades B, C, and D had significantly
  lower odds of having any cooling/air update.
- Grade B tracts had lower odds of any cooling/air update, **OR =
  exp(-0.82) = 0.44**, p = 0.005.
- Grade C tracts had lower odds of any cooling/air update, **OR =
  exp(-0.64) = 0.53**, p = 0.017.
- Grade D tracts had lower odds of any cooling/air update, **OR =
  exp(-0.74) = 0.48**, p = 0.009.
- More 2-day heatwave events were significantly associated with lower
  odds of any cooling/air update.
- Higher SVI was significantly associated with lower odds of any
  cooling/air update.
- Older median building age was significantly associated with slightly
  higher odds of any cooling/air update.

Overall, historically lower-graded areas had significantly lower odds of
having any cooling/air update compared with grade A areas, even after
accounting for heatwave exposure, social vulnerability, and building
age. This suggests that cooling-related adaptation may be less common in
historically lower-graded areas.

```{r aim3_cooling_odds}

# inferential statistics: any cooling update
 # This logistic regression tests whether the odds of having any cooling/air update differ by redlining grade.

# Outcome: any_cooling_air_update

# Predictors: dominant_redlining_grade, heatwave_events_2day, svi_overall, median_building_age
 

aim3_any_cooling_model <- glm(
  any_cooling_air_update ~
    dominant_redlining_grade +
    heatwave_events_2day +
    svi_overall +
    median_building_age,
  data = aim3_data,
  family = binomial
)

summary(aim3_any_cooling_model)


```

```{r plot_aim3}
 
# Plot results from the cooling-update logistic regression model
 
# Are tract-years more or less likely to have any residential cooling update based on redlining grade, heatwave exposure, SVI, and building age?

# Interpretation:
# OR = 1: no association
# OR > 1: higher odds of any cooling update
# OR < 1: lower odds of any cooling update
 
# Convert coefficients to odds ratios
cooling_or_data <- tidy(
  aim3_any_cooling_model,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_clean = case_when(
      term == "dominant_redlining_gradeB" ~ "Redlining grade B vs A",
      term == "dominant_redlining_gradeC" ~ "Redlining grade C vs A",
      term == "dominant_redlining_gradeD" ~ "Redlining grade D vs A",
      term == "heatwave_events_2day" ~ "2-day Heatwave Events",
      term == "svi_overall" ~ "Social Vulnerability Index",
      term == "median_building_age" ~ "Median building age",
      TRUE ~ term
    ),
    term_clean = fct_rev(term_clean)
  )

# View odds ratio table
cooling_or_data

# Plot odds ratios
cooling_or_figure <- ggplot(
  cooling_or_data,
  aes(x = estimate, y = term_clean)
) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.8
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2,
    linewidth = 0.9,
    color = "#0072B2"
  ) +
  geom_point(
    size = 3.5,
    color = "#0072B2"
  ) +
  scale_x_log10(
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("0.25", "0.5", "1.0", "2.0", "4.0")
  ) +
  labs(
    title = "Predictors of Residential Cooling Updates",
    subtitle = "Adjusted odds ratios for any cooling update; lines show 95% confidence intervals",
    x = "Adjusted odds ratio",
    y = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

cooling_or_figure

# Save for poster
ggsave(
  filename = "cooling_update_logistic_model_figure.png",
  plot = cooling_or_figure,
  width = 9,
  height = 5.5,
  dpi = 300
)
```

## Extra Models - Comparing heat, trees, and cooling by HOLC grade

In adjusted logistic regression models, census tracts in historically
lower-graded areas generally had lower odds of any residential cooling
update compared with Grade A areas.

```{r}

# Outcomes need to be transformed to binary in order to use a logistic model 
# High heatwave exposure greater than median value
# Any tree planting per census tract year
# Any residential cooling update

 
# logistic model dataset

aim3_logit_data <- tract_year_analysis %>%
  mutate(
    # Binary outcome 1: high heatwave exposure
    # Defined as above the median number of 2-day heatwave events
    high_heatwave_exposure = heatwave_events_2day >
      median(heatwave_events_2day, na.rm = TRUE),

    # Binary outcome 2: any tree planting
    any_tree_planting = trees_planted > 0,

    # Binary outcome 3: any cooling update
    # This variable already exists in your tract_year_analysis
    any_cooling_update = any_cooling_air_update,

    # Make Grade A the reference group
    dominant_redlining_grade = factor(
      dominant_redlining_grade,
      levels = c("A", "B", "C", "D")
    )
  ) %>%
  filter(
    !is.na(dominant_redlining_grade),
    !is.na(high_heatwave_exposure),
    !is.na(any_tree_planting),
    !is.na(any_cooling_update),
    !is.na(svi_overall),
    !is.na(median_building_age)
  )

 
 # Logistic model: high heatwave exposure
 
aim3_heat_logit <- glm(
  high_heatwave_exposure ~ dominant_redlining_grade + svi_overall,
  data = aim3_logit_data,
  family = binomial
)

aim3_heat_or <- tidy(
  aim3_heat_logit,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  mutate(
    outcome = "High heatwave exposure"
  )

aim3_heat_or


 # Logistic model: any tree planting
 
aim3_tree_logit <- glm(
  any_tree_planting ~ dominant_redlining_grade + svi_overall,
  data = aim3_logit_data,
  family = binomial
)

aim3_tree_or <- tidy(
  aim3_tree_logit,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  mutate(
    outcome = "Any tree planting"
  )

aim3_tree_or


 # Logistic model: any residential cooling update
 
aim3_cooling_logit <- glm(
  any_cooling_update ~ dominant_redlining_grade + svi_overall + median_building_age,
  data = aim3_logit_data,
  family = binomial
)

aim3_cooling_or <- tidy(
  aim3_cooling_logit,
  conf.int = TRUE,
  exponentiate = TRUE
) %>%
  mutate(
    outcome = "Any cooling update"
  )

aim3_cooling_or


 
# Combine redlining OR results for figure
 
aim3_or_plot_data <- bind_rows(
  aim3_heat_or,
  aim3_tree_or,
  aim3_cooling_or
) %>%
  filter(str_detect(term, "dominant_redlining_grade")) %>%
  mutate(
    redlining_grade = str_remove(term, "dominant_redlining_grade"),
    redlining_grade = case_when(
      redlining_grade == "B" ~ "B vs A",
      redlining_grade == "C" ~ "C vs A",
      redlining_grade == "D" ~ "D vs A",
      TRUE ~ redlining_grade
    ),
    redlining_grade = factor(
      redlining_grade,
      levels = c("B vs A", "C vs A", "D vs A")
    ),
    outcome = factor(
      outcome,
      levels = c(
        "High heatwave exposure",
        "Any tree planting",
        "Any cooling update"
      )
    )
  )

aim3_or_plot_data


# Plot odds ratios from aim3_or_plot_data
 
aim3_or_figure <- ggplot(
  aim3_or_plot_data,
  aes(
    x = estimate,
    y = outcome,
    color = outcome
  )
) +
  # OR = 1 means no difference from Grade A
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.8
  ) +
  # 95% confidence intervals
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.18,
    linewidth = 0.9
  ) +
  # Odds ratio point estimate
  geom_point(size = 3.5) +
  # Separate panels for each redlining comparison
  facet_wrap(~ redlining_grade, nrow = 1) +
  # Log scale is standard for odds ratios
  scale_x_log10(
    breaks = c(0.25, 0.5, 1, 2),
    labels = c("0.25", "0.5", "1.0", "2.0")
  ) +
  scale_color_manual(
    values = c(
      "High heatwave exposure" = "#D55E00",
      "Any tree planting" = "#009E73",
      "Any cooling update" = "#0072B2"
    )
  ) +
  labs(
    title = "Historical Redlining and Heat Adaptation Outcomes",
    subtitle = "Adjusted odds ratios compared with Grade A areas; lines show 95% confidence intervals",
    x = "Adjusted odds ratio compared with Grade A",
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    plot.subtitle = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

aim3_or_figure

ggsave(
  filename = "aim3_redlining_odds_ratio_figure.png",
  plot = aim3_or_figure,
  width = 13,
  height = 5.5,
  dpi = 300
)
```
