---
title: "Downloading Daymet Data"
output: html_document
date: "`r Sys.Date()`"
---

```{r set.knitr.options, include= FALSE, message=FALSE, warning=FALSE}
#-----set knitr options-----

# clear environment
rm(list = ls())

# knitr options
suppressMessages(library(knitr))
opts_chunk$set(tidy = FALSE, echo = TRUE, message = FALSE)

# Load pacman, installing if needed and other packages as needed
if (!require("pacman")) { install.packages("pacman", repos = "http://cran.r-project.org") }

# Load other packages, installing as needed.
pacman::p_load(tidyverse, dplyr, httr, jsonlite, tigris, sf, gstat, daymetr)

```

# About Daymet

Unlike NOAA (National Oceanic and Atmospheric Administration), which uses observational climate and weather data collected from ground stations, satellites, and other sources, Daymet provides gridded daily surface weather data (e.g., temperature, precipitation) for North America, derived from interpolation of ground-based station data, and is especially useful for ecological and environmental modeling where spatially continuous data are needed.

## Using Daymet R package

We will use the *daymetr* package to download weather data from 1982-2022 More information and examples of how to use the R package can be found here: <https://cran.r-project.org/web/packages/daymetr/vignettes/daymetr-vignette.html>

To download the data we need to select the locations of the estimates. Given we want to examine weather at the census tract level, we can use census tract centroids to extract estimates at these locations.

A census tract centroid is the geographic center point (latitude and longitude) of a census tract, which is a small, relatively permanent subdivision of a county used for statistical purposes. The centroid represents the approximate "middle" of the tract and is often used to link spatial data (like climate or health data) to the tract's location.

First, we need to create a CSV file with information needed for batch downloading using the *daymetr* package.

```{r centroids, echo = FALSE}

# To use Daymet we will need to make csv file with : site (GEOID), latitude, longitude

# Get 2020 census tracts for San Francisco County (FIPS: 075)
sf_tracts <- tracts(state = "CA", county = "San Francisco", year = 2020, class = "sf")

# extract centroids 
tract_centroids <- st_centroid(sf_tracts)

# select the variables 
sf_centroids <- tract_centroids %>% select(GEOID, INTPTLAT, INTPTLON)

# remove geometry 
sf_centroid_df <- st_drop_geometry(sf_centroids)

# change to numeric 
sf_centroid_df <- sf_centroid_df %>% mutate(across(c(INTPTLAT, INTPTLON), as.numeric))

# rename to use with Daymet function 
sf_centroid_df <- sf_centroid_df %>% 
                        rename(site = GEOID,
                               lat = INTPTLAT,
                               lon = INTPTLON)


# save a csv with the centroid locations for San Francisco 
write.csv(sf_centroid_df, "Data/sf_centroids.csv", row.names = FALSE)


```

### Batch downloads

The download_daymet_batch function from the daymetr R package allows users to download Daymet data for multiple locations in a single batch. Users provide a table of site information (with site names, latitude, and longitude), and the function retrieves daily weather data (e.g., temperature, precipitation) for each site over specified years. It’s useful for efficiently collecting climate data for many locations at once.

```{r daymet.download, eval = TRUE}

# warning: downloading data takes a LONG time to run, so only run once, set chunk to eval = FALSE to help prevent it from running again 

# download batch data using SF centroid locations
# make sure to indicate the year, end
# the argument "simplify" can be used to create a dataframe of the data, not using it will result your data to be in a list format 

df_batch <- download_daymet_batch(file_location = "Data/sf_centroids.csv",
start = 1982,
end = 2022,
simplify = TRUE)

sf_daymet <- df_batch 
```

### Data wrangling

We now need to fix the names of the measurement types.

```{r data.wrangling, echo = FALSE}

# what are the names of each category? use unique to find the names  
unique(sf_daymet$measurement)

sf_daymet <- sf_daymet %>%
  mutate(measurement = recode(measurement,
    "tmax..deg.c." = "tmax",
    "tmin..deg.c." = "tmin",
    "prcp..mm.day." = "prcp",
    "srad..W.m.2." = "srad",
    "swe..kg.m.2." = "swe",
    "vp..Pa." = "vp",
    "dayl..s." = "dayl"
  ))

# check names again 
unique(sf_daymet$measurement)

# check dimensions, quick check of the number of rows 
dim(sf_daymet) # 14839442

# remove the level that does not have estimates 
sf_daymet <- sf_daymet %>%
  filter(!grepl("outside DAYMET spatial coverage", measurement))

# check dimensions, quick check of the number of rows 
dim(sf_daymet) # 14839440

# check names again 
unique(sf_daymet$measurement)

# save a csv with the cleaned up data 
write.csv(sf_daymet, "Data/sf_daymet_clean.csv", row.names = FALSE)

```


# Heatwaves

According to the CDC (Centers for Disease Control and Prevention), a heatwave is a period of abnormally and uncomfortably hot and unusually humid weather, typically lasting two or more days. The definition can vary by location, but heatwaves are generally characterized by temperatures that are significantly higher than the average for a given area and time of year, and they can pose serious health risks, especially to older adults, children, and people with chronic conditions. For more information see: <https://www.cdc.gov/climate-health/php/resources/protect-yourself-from-the-dangers-of-extreme-heat.html>

Climate Change and Extreme Heat Events: <https://www.cdc.gov/climate-health/media/pdfs/ClimateChangeandExtremeHeatEvents.pdf>

Heatwave based on temperature and vapor pressure.

NCEI (National Centers for Environmental Information) calculates the 85th-percentile threshold for each MSA, both for minimum daily temperatures and for maximum daily temperatures. Using these data, EPA compared each city’s daily minimum apparent temperature records with that city’s corresponding 85th-percentile threshold. EPA identified a heat wave whenever two or more consecutive days exceeded the threshold, then quantified the following four metrics for each city and each year:

• Frequency: the number of distinct heat waves that occur every year.

• Duration: the length of each individual heat wave, in days. These data can be aggregated to find the average duration of individual heat waves over a time period such as a year or a decade.

• Season length: the number of days from the first day of the first heat wave of the year to the last day of the last heat wave, including the first and last days in the count.

• Intensity: how hot the temperature is during a heat wave, compared with the corresponding city-specific threshold. For example, if a city has an 85th-percentile threshold of 95°F, and the average of the daily minimum apparent temperatures during a three-day heat wave was 98°F, the intensity would be recorded as 3°F above the threshold.

Hence, health warnings about extreme heat are often based on NWS’s Heat Index, which is similar to apparent temperature in that it combines temperature and humidity (albeit with a different formula). This indicator specifically uses daily minimum temperature because studies show that mortality is more closely related to elevated daily minimum temperature than to daily maximum or daily mean temperature (Habeeb et al., 2015; Sarofim et al., 2016), as warm nighttime temperatures prevent the body from cooling off after a hot day. For each MSA, NCEI calculated daily maximum and minimum apparent temperature for each day based on hourly temperature and humidity measurements.

NCEI derived apparent temperature using the following equation:

AT = -1.3 + 0.92T + 2.2e

where AT is the apparent temperature (°C), T is ambient air temperature (°C), and e is water vapor pressure (kilopascals). This equation was established by Steadman (1984).

Based on Huang et al. (2021) we will examine the 97.5th percentile of apparent temperature, based on the minimum temperature. We will use the NCEI equation for apparent temperature, which does differ from Huang et al.'s (2021) equation that uses dew point instead.

Definitions:

1.  AT \> 85th percentile for 2+ days

2.  AT \> 85th percentile for 3+ days

3.  AT \> 97.5th percentile for 2+ days

4.  AT \> 97.5th percentile for 3+ days

```{r load.data, echo= FALSE}

# Step 0: Load downloaded and pre-processed clean daymet data

# sf_daymet <- read_csv("Data/sf_daymet_clean.csv")

```

Water vapor pressure vp Pa Water vapor pressure in pascals. Daily average partial pressure of water vapor.

We need to transform the vp variable from pascals to kilopascals. We divide the pressure value by 1000.

```{r apparent.temperature, echo = FALSE}
#  unique(sf_daymet$measurement)
# [1] "dayl" "prcp" "srad" "swe"  "tmax" "tmin" "vp" 

#Calculate tmean
# temp_data <- sf_daymet %>%
#   filter(measurement %in% c("tmax", "tmin")) %>%
#   pivot_wider(names_from = measurement, values_from = value) %>%
#   mutate(tmean = (tmax + tmin) / 2)


# Step 1: pivot data to wide format 
temp_data <- sf_daymet %>%
  filter(measurement %in% c("tmax", "tmin", "vp")) %>%
  pivot_wider(names_from = measurement, values_from = value) 

# kilopascals 
temp_data <- temp_data %>% 
              mutate(vp_kp = vp/1000)

# add AT 

# AT = -1.3 + 0.92T + 2.2e

temp_data <- temp_data %>% 
              mutate(AT = (-1.3 + 0.92 * (tmin) + 2.2 * (vp_kp)) )

# Reshape to long format
long_temp <- temp_data %>%
  select(site, year, yday, tmax, tmin, vp, vp_kp, AT) %>%
  pivot_longer(cols = c(tmax, tmin, vp, vp_kp, AT),
               names_to = "measurement",
               values_to = "value")

# Calculate 97.5th percentile per site & measurement

# tenyear_percentiles <- long_temp %>%
#   filter(year >= 2000 & year <= 2010) %>%     # Optional: first 10 years of data 
#   group_by(site, measurement) %>%
#   mutate(p97_5 = quantile(value, probs = 0.975, na.rm = TRUE), # percentiles are VERY similar when looking at 2000s
#             p85 = quantile(value, probs = 0.85, na.rm = TRUE))
# 


percentiles <- long_temp %>%
  filter(year >= 2000 & year <= 2020) %>%     # using 2000-2020 data in analysis 
  group_by(site, measurement) %>%
  mutate(p97_5 = quantile(value, probs = 0.975, na.rm = TRUE),
            p85 = quantile(value, probs = 0.85, na.rm = TRUE))


# filtering to only keep the AT data 
percentiles <- percentiles %>% filter(measurement == "AT")


```

How many "hot days" do we have per year per site (by year)?

```{r hot.days, echo= FALSE}

# Add threshold and flag hot days
long_temp_flagged <- percentiles %>%
  mutate(hot_day_85 = value >= p85,
         hot_day_97_5 = value >= p97_5)

# count how many days are flagged as hot days 
hot_days_summary <- long_temp_flagged %>%
  group_by(site, year) %>%
  summarise(
    n_hot_day_85 = sum(hot_day_85, na.rm = TRUE),
    n_hot_day_97_5 = sum(hot_day_97_5, na.rm = TRUE),
    .groups = "drop"
  )

# View first six rows of summary
head(hot_days_summary)


# visualizing


# ggplot(hot_days_summary, aes(x = year)) +
#   geom_line(aes(y = n_hot_day_85, color = "85th percentile")) +
#   geom_line(aes(y = n_hot_day_97_5, color = "97.5th percentile")) +
#   facet_wrap(~ site, scales = "free_y") +
#   labs(
#     title = "Number of Hot Days per Year by Site",
#     x = "Year", y = "Number of Hot Days",
#     color = "Threshold"
#   ) +
#   theme_minimal() + 
#  theme(plot.title = element_text(hjust = 0.5))




# Reshape for plotting
# hot_days_summary_long <- hot_days_summary %>%
#   pivot_longer(cols = starts_with("n_hot_day"),
#                names_to = "threshold", values_to = "count")
# 
# ggplot(hot_days_summary_long, aes(x = factor(year), y = count, fill = threshold)) +
#   geom_col(position = "dodge") +
#   facet_wrap(~ site, scales = "free_y") +
#   labs(
#     title = "Number of Hot Days per Year by Threshold",
#     x = "Year", y = "Count",
#     fill = "Threshold"
#   ) +
#   theme_minimal()  + 
#  theme(plot.title = element_text(hjust = 0.5))



# averages across city - average number of hot days per year 
hot_days_avg <- hot_days_summary %>%
  group_by(year) %>%
  summarise(
    mean_hot_day_85 = mean(n_hot_day_85, na.rm = TRUE),
    mean_hot_day_97_5 = mean(n_hot_day_97_5, na.rm = TRUE),
    .groups = "drop"
  ) 


ggplot(hot_days_avg, aes(x = year)) +
  geom_line(aes(y = mean_hot_day_85, color = "85th percentile")) +
  geom_line(aes(y = mean_hot_day_97_5, color = "97.5th percentile")) +
  labs(
    title = "Average Number of Hot Days per Year Across All SF",
    x = "Year", y = "Average Hot Days",
    color = "Threshold"
  ) +
  theme_minimal()  + 
 theme(plot.title = element_text(hjust = 0.5))


```

Now lets calculate the heat waves! We will use the four definitions we described above.

```{r heatwave.calculation, eval = FALSE}


long_temp_flagged_new <- long_temp_flagged %>%
  group_by(site, measurement, year) %>%
  arrange(yday, .by_group = TRUE) %>%
  mutate(
    # Streaks of 2+ for hot_day_85
    streak_85_2 = {
      r <- rle(hot_day_85)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 2
      ))
    },
    # Streaks of 3+ for hot_day_85
    streak_85_3 = {
      r <- rle(hot_day_85)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 3
      ))
    },
    # Streaks of 2+ for hot_day_97_5
    streak_97_5_2 = {
      r <- rle(hot_day_97_5)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 2
      ))
    },
    # Streaks of 3+ for hot_day_97_5
    streak_97_5_3 = {
      r <- rle(hot_day_97_5)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 3
      ))
    }
  ) %>%
  ungroup()

# Preview the result
head(long_temp_flagged_new)

```

```{r heatwave.counting, echo = FALSE}

# Helper function to count qualifying streaks
count_streaks <- function(x, min_length) {
  r <- rle(x)
  sum(r$values & r$lengths >= min_length)
}

# Main heat wave summary
heat_wave_summary <- long_temp_flagged %>%
  filter(measurement == "AT") %>%
  group_by(site, year) %>%
  arrange(yday, .by_group = TRUE) %>%  # ensure correct order
  summarise(
    hw_2day_85   = count_streaks(hot_day_85, 2),
    hw_3day_85   = count_streaks(hot_day_85, 3),
    hw_2day_97_5 = count_streaks(hot_day_97_5, 2),
    hw_3day_97_5 = count_streaks(hot_day_97_5, 3),
    .groups = "drop"
  )

# save a copy of the heat wave data 

# write.csv(heat_wave_summary, "Data/heat_wave_summary.csv", row.names = FALSE)

  
```
