---
title: "Final Combined Dataset"
author: "Nancy Carmona"
date: "`r Sys.Date()`"
output: html_document
---
 
```{r setup, include=FALSE}
#-----setup options-----

# clear the workspace
rm(list = ls())

# knitr options, shows the code in the knitted HTML file 
knitr::opts_chunk$set(echo = TRUE)

```

```{r load_libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# load pacman, installing if needed
if (!require("pacman")) {install.packages("pacman")}

# load other packages, installing as needed
pacman::p_load(tidyverse, lubridate, stringr, tigris, sf, leaflet, readr)

```


# Introduction

This code combines several datasets to create a final team data file for all to work on their individual research questions. This includes: 

* Building permits (includes "cooling" indicator, year built and building age [in relation to 2022])

* Heatwave indicators (defined as 2+ days above 95th percentile of historical normal [30-year avg])​

* CDC SVI (main themes and overall index)​

* Historical redlining grades 

* Trees planted (trees planted from 2012 – 2022)


Each dataset is cleaned separately first. Then we join them together by census tract and year.

```{r load_data}
# load cleaned data sets for merging final data 

# Building permits
final_building_data <- read.csv("Data/final_building_data.csv")

# Heatwave indicators 
sf_tracts_hw <- read.csv("Data/heatwaves_2012_2022.csv")

# SVI Census Tract level 
#census level data 
svi_census <- read.csv("Data/svi_2022_census.csv")

# SVI Zip Code level 
svi_zcta <- read.csv("Data/svi_2022_zcta.csv")

# SF Redlining 
redlining <- read_sf("Data/mappinginequality.json")

# Planted trees
sf_tree <- read.csv("Data/sf_tree.csv")

```

## Building Data 

This code uses final building data that was created using "Building Data Code.Rmd".

The file "final_building_data.csv" is also found in the Box Folder: PINC 2026 > Data > Building Permit Data Sets

```{r data_wrangle_building}

# Clean building permit data for yearly analysis

building_year_clean <- final_building_data %>%
  mutate(
    census_tract_geoid = as.character(round(census_tract_geoid, 0)),
    zipcode = as.character(zipcode),
    neighborhood = as.character(neighborhood),
    year = year(ymd_hms(filed_date)),
    building_age = 2022 - year_built # age as of 2022? 
  ) %>%
  filter(
    !is.na(year),
    year >= 2012,
    year <= 2022
  )

# Check cleaned building data
building_year_clean %>%
  select(
    parcel_id,
    census_tract_geoid,
    zipcode,
    neighborhood,
    year,
    cooling_air_update
  ) %>%
  head()

# Summarize building permits to census tract + year

building_tract_year <- building_year_clean %>%
  filter(!is.na(census_tract_geoid)) %>%
  group_by(census_tract_geoid, year) %>%
  summarise(
    total_permits = n(),
    cooling_air_updates = sum(cooling_air_update, na.rm = TRUE),
    percent_cooling_air = round(
      cooling_air_updates / total_permits * 100,
      2
    ),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE),
    .groups = "drop"
  )

# Check result
building_tract_year %>%
  head()

```

## Heat Data 

This code uses final heat data that was created using "Downloading_Daymet_Data.Rmd".

The file "heatwaves_2012_2022.csv" is also found in the Box Folder: PINC 2026 > Data > Daymet Data

```{r data_wrangle_heat}

# Clean and summarize heatwave data to census tract + year

heat_tract_year <- sf_tracts_hw %>%
  mutate(
    census_tract_geoid = as.character(site),
    year = as.numeric(year)
  ) %>%
  filter(
    year >= 2012,
    year <= 2022
  ) %>%
  group_by(census_tract_geoid, year) %>%
  summarise(
    heat_days_95 = sum(hw95_day, na.rm = TRUE),
    heatwave_days_2day = sum(hw95_2day, na.rm = TRUE),
    heatwave_days_3day = sum(hw95_3day, na.rm = TRUE),
    heatwave_events_2day = n_distinct(hw_event_id[hw95_2day == TRUE]),
    heatwave_events_3day = n_distinct(hw_event_id[hw95_3day == TRUE]),
    mean_apparent_temp = mean(AT, na.rm = TRUE),
    .groups = "drop"
  )

# Check result
heat_tract_year %>%
  head()

```

## SVI Data 

This code uses SVI data that was created using "Downloading_SVI_Data.Rmd". The SVI data was downloaded for just CA (not the whole US).

The file "svi_2022_census.csv" and "svi_2022_zcta.csv" is also found in the Box Folder: PINC 2026 > Data > SVI Data Sets

```{r data_wrangle_svi}

# Clean census tract-level SVI data

svi_tract <- svi_census %>%
  mutate(
    census_tract_geoid = as.character(round(site, 0))
  ) %>%
  select(
    census_tract_geoid,
    svi_overall = RPL_THEMES,
    svi_socioeconomic = RPL_THEME1,
    svi_household = RPL_THEME2,
    svi_race_ethnicity = RPL_THEME3,
    svi_housing_transportation = RPL_THEME4
  )


# Clean ZIP/ZCTA-level SVI data

svi_zip <- svi_zcta %>%
  mutate(
    zipcode = as.character(FIPS)
  ) %>%
  select(
    zipcode,
    zip_svi_overall = RPL_THEMES,
    zip_svi_socioeconomic = RPL_THEME1,
    zip_svi_household = RPL_THEME2,
    zip_svi_race_ethnicity = RPL_THEME3,
    zip_svi_housing_transportation = RPL_THEME4
  )


# Check cleaned SVI data

glimpse(svi_tract)
glimpse(svi_zip)

```

## Redlining Data 

This code uses redlining data downloaded from the Mapping Inequality Webiste. The data is also used in the "Redlining.Rmd" which process the data to align with SF.

The file "mappinginequality.json" is also found in the Box Folder: PINC 2026 > Data > Redlining Data


```{r data_wrangle_redlining}

# Clean historical redlining data for San Francisco

sf_redlining_clean <- redlining %>%
  filter(
    city == "San Francisco",
    state == "CA"
  ) %>%
  st_transform(4326) %>%
  st_make_valid() %>%
  mutate(
    redlining_area_id = area_id,
    redlining_grade = grade,
    redlining_category = category,
    redlining_label = label
  ) %>%
  select(
    redlining_area_id,
    redlining_grade,
    redlining_category,
    redlining_label,
    geometry
  )

# Check cleaned redlining data
sf_redlining_clean %>%
  st_drop_geometry() %>%
  count(redlining_grade, redlining_category, sort = TRUE)

```

## Tree Data 

This code uses final tree data that was created using "Tree Canopy.Rmd". 

The file "sf_tree.csv" is also found in the Box Folder: PINC 2026 > Data > Tree Canopy Data

```{r data_wrangle_trees}
# Clean simplified tree data

tree_clean <- sf_tree %>%
  st_drop_geometry() %>%
  mutate(
    census_tract_geoid = as.character(GEOID),
    year = as.numeric(Year)
  ) %>%
  filter(
    !is.na(census_tract_geoid),
    !is.na(year),
    year >= 2012,
    year <= 2022
  ) %>%
  select(
    tree_id = TreeID,
    census_tract_geoid,
    year
  )

# Check cleaned tree data
glimpse(tree_clean)

# Count trees planted by census tract and year

tree_tract_year <- tree_clean %>%
  group_by(census_tract_geoid, year) %>%
  summarise(
    trees_planted = n(),
    .groups = "drop"
  )

# Check tree tract-year data
tree_tract_year %>%
  head()


# Check tree data summary

tree_tract_year %>%
  summarise(
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE),
    total_tract_years = n(),
    total_trees_planted = sum(trees_planted, na.rm = TRUE)
  )

```

## Creating Final Dataset
 
```{r standardize_tract_ids}
# Make sure all census tract IDs use the same 11-digit format

building_tract_year <- building_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0"),
    year = as.numeric(year)
  )

heat_tract_year <- heat_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0"),
    year = as.numeric(year)
  )

tree_tract_year <- tree_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0"),
    year = as.numeric(year)
  )

svi_tract <- svi_tract %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0")
  )

svi_zip <- svi_zip %>%
  mutate(
    zipcode = as.character(zipcode)
  )
```



```{r prepare_redlining_tract}
# Prepare redlining data at the census tract level
 
options(tigris_use_cache = TRUE)

# Get San Francisco census tract boundaries
sf_tracts_for_join <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2020,
  cb = TRUE
) %>%
  st_transform(4326) %>%
  select(
    census_tract_geoid = GEOID,
    geometry
  )

# Make sure redlining polygons are clean
sf_redlining_clean <- sf_redlining_clean %>%
  st_transform(4326) %>%
  st_make_valid()

# Turn off s2 for polygon intersection
sf_use_s2(FALSE)

# Intersect census tracts with redlining areas
redlining_tract_overlap <- st_intersection(
  sf_tracts_for_join,
  sf_redlining_clean
) %>%
  st_transform(3310) %>%
  mutate(
    overlap_area = as.numeric(st_area(geometry))
  ) %>%
  st_drop_geometry()

sf_use_s2(TRUE)

# Keep the redlining grade with the largest overlap in each tract
redlining_tract <- redlining_tract_overlap %>%
  group_by(census_tract_geoid) %>%
  slice_max(overlap_area, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    census_tract_geoid,
    dominant_redlining_grade = redlining_grade,
    dominant_redlining_category = redlining_category,
    dominant_redlining_area_id = redlining_area_id
  )

# Check result
redlining_tract %>%
  count(dominant_redlining_grade, dominant_redlining_category, sort = TRUE)

```

```{r create_tract_year_full}
# Create the full census tract-year dataset

tract_year_full <- building_tract_year %>%
  full_join(
    heat_tract_year,
    by = c("census_tract_geoid", "year")
  ) %>%
  full_join(
    tree_tract_year,
    by = c("census_tract_geoid", "year")
  ) %>%
  left_join(
    svi_tract,
    by = "census_tract_geoid"
  ) %>%
  left_join(
    redlining_tract,
    by = "census_tract_geoid"
  )

# Check final tract-year dataset
glimpse(tract_year_full)

```



Create tract-to-ZIP and tract-to-neighborhood lookup tables

```{r create_zip_neighborhood_lookup}
# Create lookup tables from census tracts to ZIP codes and neighborhoods

# For each census tract, keep the most common ZIP code
tract_zip_lookup <- final_building_data %>%
  mutate(
    census_tract_geoid = str_pad(
      as.character(round(census_tract_geoid, 0)),
      width = 11,
      side = "left",
      pad = "0"
    ),
    zipcode = as.character(zipcode)
  ) %>%
  filter(
    !is.na(census_tract_geoid),
    !is.na(zipcode)
  ) %>%
  count(census_tract_geoid, zipcode, sort = TRUE) %>%
  group_by(census_tract_geoid) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    census_tract_geoid,
    zipcode
  )


# For each census tract, keep the most common neighborhood
tract_neighborhood_lookup <- final_building_data %>%
  mutate(
    census_tract_geoid = str_pad(
      as.character(round(census_tract_geoid, 0)),
      width = 11,
      side = "left",
      pad = "0"
    ),
    neighborhood = as.character(neighborhood)
  ) %>%
  filter(
    !is.na(census_tract_geoid),
    !is.na(neighborhood),
    neighborhood != ""
  ) %>%
  count(census_tract_geoid, neighborhood, sort = TRUE) %>%
  group_by(census_tract_geoid) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    census_tract_geoid,
    neighborhood
  )

# Check data

tract_zip_lookup %>%
  head()

tract_neighborhood_lookup %>%
  head()
```


Create ZIP code-year dataset

This summarizes the tract-year dataset up to ZIP code and year.


```{r create_zip_year_full}
# Create ZIP code-year dataset

zip_year_full <- tract_year_full %>%
  left_join(
    tract_zip_lookup,
    by = "census_tract_geoid"
  ) %>%
  filter(!is.na(zipcode)) %>%
  group_by(zipcode, year) %>%
  summarise(
    total_permits = sum(total_permits, na.rm = TRUE),
    cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(
      cooling_air_updates / total_permits * 100,
      2
    ),

    median_year_built = median(median_year_built, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),

    heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    heatwave_days_2day = mean(heatwave_days_2day, na.rm = TRUE),
    heatwave_days_3day = mean(heatwave_days_3day, na.rm = TRUE),
    heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    heatwave_events_3day = mean(heatwave_events_3day, na.rm = TRUE),
    mean_apparent_temp = mean(mean_apparent_temp, na.rm = TRUE),

    trees_planted = sum(trees_planted, na.rm = TRUE),

    mean_svi_overall = mean(svi_overall, na.rm = TRUE),
    mean_svi_socioeconomic = mean(svi_socioeconomic, na.rm = TRUE),
    mean_svi_household = mean(svi_household, na.rm = TRUE),
    mean_svi_race_ethnicity = mean(svi_race_ethnicity, na.rm = TRUE),
    mean_svi_housing_transportation = mean(svi_housing_transportation, na.rm = TRUE),

    .groups = "drop"
  ) %>%
  left_join(
    svi_zip,
    by = "zipcode"
  )

# Check ZIP code-year dataset
glimpse(zip_year_full)

```

Create neighborhood-year dataset

```{r create_neighborhood_year_full}
# Create neighborhood-year dataset

neighborhood_year_full <- tract_year_full %>%
  left_join(
    tract_neighborhood_lookup,
    by = "census_tract_geoid"
  ) %>%
  filter(!is.na(neighborhood)) %>%
  group_by(neighborhood, year) %>%
  summarise(
    total_permits = sum(total_permits, na.rm = TRUE),
    cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(
      cooling_air_updates / total_permits * 100,
      2
    ),

    median_year_built = median(median_year_built, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),

    heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    heatwave_days_2day = mean(heatwave_days_2day, na.rm = TRUE),
    heatwave_days_3day = mean(heatwave_days_3day, na.rm = TRUE),
    heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    heatwave_events_3day = mean(heatwave_events_3day, na.rm = TRUE),
    mean_apparent_temp = mean(mean_apparent_temp, na.rm = TRUE),

    trees_planted = sum(trees_planted, na.rm = TRUE),

    mean_svi_overall = mean(svi_overall, na.rm = TRUE),
    mean_svi_socioeconomic = mean(svi_socioeconomic, na.rm = TRUE),
    mean_svi_household = mean(svi_household, na.rm = TRUE),
    mean_svi_race_ethnicity = mean(svi_race_ethnicity, na.rm = TRUE),
    mean_svi_housing_transportation = mean(svi_housing_transportation, na.rm = TRUE),

    .groups = "drop"
  )

# Check neighborhood-year dataset
glimpse(neighborhood_year_full)

```

Check final datasets

```{r check_final_geography_datasets}
# Check number of rows and years in each dataset

tract_year_full %>%
  summarise(
    geography = "Census tract",
    rows = n(),
    geographies = n_distinct(census_tract_geoid),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

zip_year_full %>%
  summarise(
    geography = "ZIP code",
    rows = n(),
    geographies = n_distinct(zipcode),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

neighborhood_year_full %>%
  summarise(
    geography = "Neighborhood",
    rows = n(),
    geographies = n_distinct(neighborhood),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )


```


```{r save_final_datasets}

# Save final  datasets
write_csv(tract_year_full, "Data/tract_year_full.csv")
write_csv(zip_year_full, "Data/zip_year_full.csv")
write_csv(neighborhood_year_full, "Data/neighborhood_year_full.csv")
