---
title: "Final Combined Dataset"
author: "Nancy Carmona"
date: "`r Sys.Date()`"
output: html_document
---
 
```{r setup, include=FALSE}
#-----setup options-----

# clear the workspace
rm(list = ls())

# knitr options, shows the code in the knitted HTML file 
knitr::opts_chunk$set(echo = TRUE)

```

```{r load_libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# load pacman, installing if needed
if (!require("pacman")) {install.packages("pacman")}

# load other packages, installing as needed
pacman::p_load(tidyverse, lubridate, stringr, tigris, sf, leaflet, readr, viridis)
 
```


# Introduction

This code combines several datasets to create a final team data file for all to work on their individual research questions. This includes: 

* Building permits (includes "cooling" indicator, year built and building age [in relation to 2022])

* Heatwave indicators (defined as 2+ days above 95th percentile of historical normal [30-year avg])​

* CDC SVI (main themes and overall index)​

* Historical redlining grades 

* Trees planted (trees planted from 2012 – 2022)


Each dataset is cleaned separately first. Then we join them together by census tract and year.

```{r load_data}
# load cleaned data sets for merging final data 

# Building permits
final_building_data <- read.csv("Data/final_building_data.csv")

# Heatwave indicators 
# updated on 7/2 with 2013 to 2022
sf_tracts_hw <- read.csv("Data/heatwaves_2013_2022.csv")

# SVI Census Tract level 
#census level data 
svi_census <- read.csv("Data/svi_2022_census.csv")

# SVI Zip Code level 
svi_zcta <- read.csv("Data/svi_2022_zcta.csv")

# SF Redlining 
redlining <- read_sf("Data/mappinginequality.json")

# Planted trees
sf_tree <- read.csv("Data/sf_tree.csv")

```

## Building Data 

This code uses final building data that was created using "Building Data Code.Rmd".

The file "final_building_data.csv" is also found in the Box Folder: PINC 2026 > Data > Building Permit Data Sets

Note: Filter out building updates not related to peoples homes (e.g., commercial, industrial, government)

Building permit records were restricted to single-family and multi-family residential properties to better capture cooling upgrades in places where people live, rather than commercial or institutional buildings.


```{r residential_subset}

# residential only permits 
 final_building_residential <- final_building_data %>%
   filter(property_use %in% c(
     "Single Family Residential",
     "Multi-Family Residential"
   ))
 
```

```{r data_wrangle_building}


# Clean building permit data for yearly analysis

building_year_clean <- final_building_residential %>%
  mutate(
    census_tract_geoid = as.character(round(census_tract_geoid, 0)),
    zipcode = as.character(zipcode),
    neighborhood = as.character(neighborhood),
    year = year(ymd_hms(filed_date)),
    building_age = 2022 - year_built # age as of 2022? 
  ) %>%
  filter(
    !is.na(year),
    year >= 2013, # changed min to 2013 
    year <= 2022
  )

# Check cleaned building data
building_year_clean %>%
  select(
    parcel_id,
    census_tract_geoid,
    zipcode,
    neighborhood,
    year,
    cooling_air_update
  ) %>%
  head()

# Summarize building permits to census tract + year

building_tract_year <- building_year_clean %>%
  filter(!is.na(census_tract_geoid)) %>%
  group_by(census_tract_geoid, year) %>%
  summarise(
    total_permits = n(),
    cooling_air_updates = sum(cooling_air_update, na.rm = TRUE),
    percent_cooling_air = round(
      cooling_air_updates / total_permits * 100,
      2
    ),
    median_year_built = median(year_built, na.rm = TRUE),
    median_building_age = median(building_age, na.rm = TRUE),
    .groups = "drop"
  )

# Check result
building_tract_year %>%
  head()

```

### Cooling Upgrades Map 

One tract had many repeated cooling permits concentrated in a few parcels! So the map uses categories so we can see the general pattern. 2015 had some outliers from normal ranges.. again something happened in 2015?

```{r map_residential_cooling}

# Map total residential cooling updates by census tract
# with corrected GEOIDs and cropped SF map extent

options(tigris_use_cache = TRUE)

# Summarize total residential cooling updates by tract
building_map_data <- building_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(
      as.character(census_tract_geoid),
      width = 11,
      side = "left",
      pad = "0"
    )
  ) %>%
  group_by(census_tract_geoid) %>%
  summarise(
    total_residential_cooling_updates = sum(cooling_air_updates, na.rm = TRUE),
    .groups = "drop"
  )

# Get San Francisco census tract boundaries from tigris
sf_tracts <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2020,
  cb = TRUE
) %>%
  st_transform(4326) %>%
  select(
    census_tract_geoid = GEOID,
    geometry
  )

# Join cooling data to tract boundaries
building_map_sf <- sf_tracts %>%
  left_join(building_map_data, by = "census_tract_geoid") %>%
  mutate(
    total_residential_cooling_updates = replace_na(
      total_residential_cooling_updates,
      0
    ),
    cooling_update_category = cut(
      total_residential_cooling_updates,
      breaks = c(-Inf, 0, 1, 2, 5, 10, 25, 50, Inf),
      labels = c("0", "1", "2", "3–5", "6–10", "11–25", "26–50", "51+")
    )
  )

# Crop map extent to mainland San Francisco
sf_bbox <- st_bbox(
  c(
    xmin = -122.53,
    xmax = -122.35,
    ymin = 37.70,
    ymax = 37.84
  ),
  crs = st_crs(building_map_sf)
)

building_map_sf_crop <- st_crop(building_map_sf, sf_bbox)

# Map total residential cooling updates
ggplot(building_map_sf_crop) +
  geom_sf(
    aes(fill = cooling_update_category),
    color = "white",
    linewidth = 0.1
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    na.value = "grey90",
    name = "Cooling\nupdates"
  ) +
  labs(
    title = "Residential Cooling Updates by Census Tract",
    subtitle = "Total residential cooling-related permits, 2013–2022"
  ) +
  coord_sf(
    xlim = c(-122.53, -122.35),
    ylim = c(37.70, 37.84),
    datum = NA,
    expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5)
  )

```


```{r}
# Neighborhood-level map of residential cooling upgrades
# Uses local shapefile folder: "Neighborhood Shapefile"
 

#  Read SF neighborhood shapefile
sf_neighborhoods <- st_read(
  "Data/Neighborhood Shapefile/geo_export_c03b1232-1f9b-41f4-9f16-5fc136f98717.shp",
  quiet = TRUE
) %>%
  st_transform(4326)

#  Rename neighborhood column
# If this line errors, run names(sf_neighborhoods) and replace nhood
# with the actual neighborhood name column.
sf_neighborhoods <- sf_neighborhoods %>%
  rename(neighborhood = nhood) %>%
  mutate(
    neighborhood_clean = str_to_lower(neighborhood),
    neighborhood_clean = str_squish(neighborhood_clean)
  )

#  Summarize residential cooling upgrades by neighborhood
cooling_neighborhood_data <- building_year_clean %>%
  filter(!is.na(neighborhood)) %>%
  mutate(
    neighborhood_clean = str_to_lower(neighborhood),
    neighborhood_clean = str_squish(neighborhood_clean)
  ) %>%
  group_by(neighborhood_clean) %>%
  summarise(
    total_residential_permits = n(),
    total_residential_cooling_updates = sum(cooling_air_update == 1, na.rm = TRUE),
    percent_residential_cooling = if_else(
      total_residential_permits > 0,
      round(total_residential_cooling_updates / total_residential_permits * 100, 2),
      NA_real_
    ),
    .groups = "drop"
  )

# Join data to neighborhood boundaries and create bins
cooling_neighborhood_sf <- sf_neighborhoods %>%
  left_join(
    cooling_neighborhood_data,
    by = "neighborhood_clean"
  ) %>%
  mutate(
    total_residential_cooling_updates = replace_na(total_residential_cooling_updates, 0),
    cooling_update_category = cut(
      total_residential_cooling_updates,
      breaks = c(-Inf, 0, 5, 10, 25, 50, 100, Inf),
      labels = c("0", "1–5", "6–10", "11–25", "26–50", "51–100", "101+")
    )
  )

# Map residential cooling upgrades by neighborhood
ggplot(cooling_neighborhood_sf) +
  geom_sf(
    aes(fill = cooling_update_category),
    color = "white",
    linewidth = 0.25
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    na.value = "grey90",
    name = "Cooling\nupdates"
  ) +
  labs(
    title = "Residential Cooling Updates by Neighborhood",
    subtitle = "Total residential cooling-related permits, 2013–2022"
  ) +
  coord_sf(datum = NA, expand = FALSE) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5)
  )
```



## Heat Data 

This code uses final heat data that was created using "Downloading_Daymet_Data.Rmd".

The file "heatwaves_2012_2022.csv" is also found in the Box Folder: PINC 2026 > Data > Daymet Data

```{r data_wrangle_heat}

# Clean and summarize heat data to census tract + year
# Noted updated 7/2 with exposure period of 2013-2022

heat_tract_year <- sf_tracts_hw %>%
  mutate(
    census_tract_geoid = as.character(site),
    date = as.Date(date),
    year = as.numeric(year)
  ) %>%
  filter(
    year >= 2013,
    year <= 2022
  ) %>%
  group_by(census_tract_geoid, year) %>%
  summarise(
    # Extreme heat days: any day with AT >= tract-specific 95th percentile
    heat_days_95 = sum(hw95_day, na.rm = TRUE),
    
    # Heatwave days: days that are part of multi-day heatwave events
    heatwave_days_2day = sum(hw95_2day, na.rm = TRUE),
    heatwave_days_3day = sum(hw95_3day, na.rm = TRUE),
    
    # Heatwave events: number of distinct multi-day events
    heatwave_events_2day = n_distinct(hw_event_id[hw95_2day == TRUE]),
    heatwave_events_3day = n_distinct(hw_event_id[hw95_3day == TRUE]),
    
    # Average apparent temperature across all days in that tract-year
    mean_apparent_temp = mean(AT, na.rm = TRUE),
    
    .groups = "drop"
  )

# Check result
heat_tract_year %>%
  head()

```


### Heatwaves Map 


```{r}
# Map mean annual 2-day heatwave events by census tract
# Uses existing sf_tracts and sf_bbox from the cooling map code

heat_map_data <- heat_tract_year %>%
  mutate(
    census_tract_geoid = stringr::str_pad(
      as.character(census_tract_geoid),
      width = 11,
      side = "left",
      pad = "0"
    )
  ) %>%
  group_by(census_tract_geoid) %>%
  summarise(
    mean_heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    .groups = "drop"
  )

heat_map_sf <- sf_tracts %>%
  left_join(
    heat_map_data,
    by = "census_tract_geoid"
  )

heat_map_sf_crop <- st_crop(heat_map_sf, sf_bbox)

ggplot(heat_map_sf_crop) +
  geom_sf(
    aes(fill = mean_heatwave_events_2day),
    color = "white",
    linewidth = 0.1
  ) +
  scale_fill_viridis_c(
    option = "inferno",
    na.value = "grey90",
    name = "Mean annual\nheatwave events"
  ) +
  labs(
    title = "Heatwave Events by Census Tract",
    subtitle = "Average annual 2-day heatwave events, 2013–2022"
  ) +
  coord_sf(
    xlim = c(-122.53, -122.35),
    ylim = c(37.70, 37.84),
    datum = NA,
    expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5)
  )
```


## SVI Data 

This code uses SVI data that was created using "Downloading_SVI_Data.Rmd". The SVI data was downloaded for just CA (not the whole US).

The file "svi_2022_census.csv" and "svi_2022_zcta.csv" is also found in the Box Folder: PINC 2026 > Data > SVI Data Sets

```{r data_wrangle_svi}

# Clean census tract-level SVI data

svi_tract <- svi_census %>%
  mutate(
    census_tract_geoid = as.character(round(site, 0))
  ) %>%
  select(
    census_tract_geoid,
    svi_overall = RPL_THEMES,
    svi_socioeconomic = RPL_THEME1,
    svi_household = RPL_THEME2,
    svi_race_ethnicity = RPL_THEME3,
    svi_housing_transportation = RPL_THEME4
  )


# Clean ZIP/ZCTA-level SVI data

svi_zip <- svi_zcta %>%
  mutate(
    zipcode = as.character(FIPS)
  ) %>%
  select(
    zipcode,
    zip_svi_overall = RPL_THEMES,
    zip_svi_socioeconomic = RPL_THEME1,
    zip_svi_household = RPL_THEME2,
    zip_svi_race_ethnicity = RPL_THEME3,
    zip_svi_housing_transportation = RPL_THEME4
  )


# Check cleaned SVI data

glimpse(svi_tract)
glimpse(svi_zip)

```
### SVI Map 

```{r, echo = FALSE}
 
# fix tract variable format 
svi_tract <- svi_tract %>%
  mutate(
    census_tract_geoid = stringr::str_pad(
      as.character(census_tract_geoid),
      width = 11,
      side = "left",
      pad = "0"
    ) )

# load shape outline 
options(tigris_use_cache = TRUE)

# Load San Francisco census tracts
sf_tracts <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2020, # 2020 shape 
  cb = TRUE,
  class = "sf"
)

# Join SVI data to tract shapes
svi_map_sf <- sf_tracts %>%
  mutate(
    census_tract_geoid = GEOID
  ) %>%
  left_join(
    svi_tract,
    by = "census_tract_geoid"
  )

# Create city boundary from tract polygons
sf_boundary <- svi_map_sf %>%
  summarise(geometry = st_union(geometry))

# Map overall SVI
ggplot() +
  geom_sf(
    data = svi_map_sf,
    aes(fill = svi_overall),
    color = "white",
    linewidth = 0.1
  ) +
  geom_sf(
    data = sf_boundary,
    fill = NA,
    color = "black",
    linewidth = 0.6
  ) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "grey90",
    name = "Overall\nSVI"
  ) +
  labs(
    title = "Social Vulnerability Index by Census Tract",
  ) +
  coord_sf(
    xlim = c(-122.53, -122.35),
    ylim = c(37.70, 37.84),
    datum = NA,
    expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

## Redlining Data 

This code uses redlining data downloaded from the Mapping Inequality Webiste. The data is also used in the "Redlining.Rmd" which process the data to align with SF.

The file "mappinginequality.json" is also found in the Box Folder: PINC 2026 > Data > Redlining Data


```{r data_wrangle_redlining}

# Clean historical redlining data for San Francisco

sf_redlining_clean <- redlining %>%
  filter(
    city == "San Francisco",
    state == "CA"
  ) %>%
  st_transform(4326) %>%
  st_make_valid() %>%
  mutate(
    redlining_area_id = area_id,
    redlining_grade = grade,
    redlining_category = category,
    redlining_label = label
  ) %>%
  select(
    redlining_area_id,
    redlining_grade,
    redlining_category,
    redlining_label,
    geometry
  )

# Check cleaned redlining data
sf_redlining_clean %>%
  st_drop_geometry() %>%
  count(redlining_grade, redlining_category, sort = TRUE)

```

### Redlined Map 

```{r map_redlining, echo = FALSE}

# use shape of SF

options(tigris_use_cache = TRUE)

# Load San Francisco census tracts
sf_tracts <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2022,
  cb = TRUE,
  class = "sf"
)

# Make sure tracts use the same CRS as your redlining layer
sf_tracts <- st_transform(sf_tracts, st_crs(sf_redlining))

# Create city boundary from tract polygons
sf_boundary <- sf_tracts %>%
  summarise(geometry = st_union(geometry))

# HOLC colors
holc_colors <- c(
  "A" = alpha("green", 0.6),
  "B" = alpha("blue", 0.6),
  "C" = alpha("yellow", 0.6),
  "D" = alpha("red", 0.6)
)

# Map redlining with census tract and city outlines
redlining_map <- ggplot() +
  geom_sf(
    data = sf_redlining,
    aes(fill = grade),
    color = "white",
    linewidth = 0.15
  ) +
  geom_sf(
    data = sf_tracts,
    fill = NA,
    color = "grey35",
    linewidth = 0.12
  ) +
  geom_sf(
    data = sf_boundary,
    fill = NA,
    color = "grey1",
    linewidth = 0.6
  ) +
  scale_fill_manual(
    values = holc_colors,
    na.value = "grey90",
    name = "HOLC Grade"
  ) +
  labs(
    title = "Historical Redlining in San Francisco",
    subtitle = "HOLC Grades"
  ) +
  coord_sf(
    xlim = c(-122.53, -122.35),
    ylim = c(37.70, 37.84),
    datum = NA,
    expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
    
  )

 
redlining_map

```



## Tree Data 

This code uses final tree data that was created using "Tree Canopy.Rmd". 

The file "sf_tree.csv" is also found in the Box Folder: PINC 2026 > Data > Tree Canopy Data

```{r data_wrangle_trees}
# Clean simplified tree data

tree_clean <- sf_tree %>%
  st_drop_geometry() %>%
  mutate(
    census_tract_geoid = as.character(GEOID),
    year = as.numeric(Year)
  ) %>%
  filter(
    !is.na(census_tract_geoid),
    !is.na(year),
    year >= 2013,
    year <= 2022
  ) %>%
  select(
    tree_id = TreeID,
    census_tract_geoid,
    year
  )

# Check cleaned tree data
glimpse(tree_clean)

# Count trees planted by census tract and year

tree_tract_year <- tree_clean %>%
  group_by(census_tract_geoid, year) %>%
  summarise(
    trees_planted = n(),
    .groups = "drop"
  )

# Check tree tract-year data
tree_tract_year %>%
  head()


# Check tree data summary

tree_tract_year %>%
  summarise(
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE),
    total_tract_years = n(),
    total_trees_planted = sum(trees_planted, na.rm = TRUE)
  )

```

### Tree Map

Tree planting counts were highly skewed! So we mapped categories instead of a continuous scale to keep one high-count tract from not showing patterns across city. But scale can be misleading?

```{r}
#use geography

options(tigris_use_cache = TRUE)

# Count total trees planted by census tract
tree_tract_total <- sf_tree %>%
    filter(Year != 2012) %>%
  mutate(
    GEOID = as.character(GEOID),
    GEOID = stringr::str_pad(GEOID, width = 11, pad = "0")
  ) %>%
  group_by(GEOID) %>%
  summarise(
    total_trees_planted = n(),
    .groups = "drop"
  )

# Get San Francisco census tract boundaries
sf_tracts <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2020,
  cb = TRUE,
  class = "sf"
) %>%
  mutate(GEOID = as.character(GEOID))

# Join tree counts to census tract boundaries
sf_tree_map <- sf_tracts %>%
  left_join(tree_tract_total, by = "GEOID") %>%
  mutate(
    total_trees_planted = replace_na(total_trees_planted, 0),

    # Bins for total trees planted
    trees_bin = case_when(
      total_trees_planted == 0 ~ "0",
      total_trees_planted >= 1   & total_trees_planted <= 5   ~ "1–5",
      total_trees_planted >= 6   & total_trees_planted <= 10  ~ "6–10",
      total_trees_planted >= 11  & total_trees_planted <= 25  ~ "11–25",
      total_trees_planted >= 26  & total_trees_planted <= 50  ~ "26–50",
      total_trees_planted >= 51  & total_trees_planted <= 100 ~ "51–100",
      total_trees_planted >= 101 & total_trees_planted <= 200 ~ "101–200",
      total_trees_planted >= 201 ~ "201+"
    ),
    trees_bin = factor(
      trees_bin,
      levels = c("0", "1–5", "6–10", "11–25", "26–50", "51–100", "101–200", "201+")
    )
  )

# Remove tiny offshore geometries/islands to trim the map
sf_tree_map_main <- sf_tree_map %>%
  mutate(area = as.numeric(st_area(.))) %>%
  filter(area > quantile(area, 0.02, na.rm = TRUE))

# 5. Create tight bounding box
sf_bbox <- st_bbox(sf_tree_map_main)

# 6. Make final map
sf_tree_bin_map <- ggplot(sf_tree_map) +
  geom_sf(
    aes(fill = trees_bin),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    name = "Trees\nplanted",
    drop = FALSE,
    na.value = "grey90"
  ) +
  coord_sf(
    xlim = c(-122.53, -122.35),
    ylim = c(37.70, 37.84),
    datum = NA,
    expand = FALSE
  ) +
  labs(
    title = "Trees Planted by Census Tract",
    subtitle = "Total trees planted, 2013–2022"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18),
    plot.subtitle = element_text(size = 13),
    plot.margin = margin(10, 10, 10, 10),
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )

# Print map
sf_tree_bin_map


# Save map!
ggsave(
  "sf_trees_by_census_tract_binned.png",
  plot = sf_tree_bin_map,
  width = 9,
  height = 6,
  dpi = 300
)
```

```{r}
 
sf_boundary <- sf_tree_map %>%
  summarise(geometry = st_union(geometry))

ggplot() +
  geom_sf(
    data = sf_tree_map,
    aes(fill = total_trees_planted),
    color = "white",
    linewidth = 0.12
  ) +
  geom_sf(
    data = sf_boundary,
    fill = NA,
    color = "black",
    linewidth = 0.5
  ) +
  scale_fill_viridis_c(
    option = "mako",
    direction = -1,
    na.value = "grey90",
    name = "Trees\nplanted"
  ) +
  labs(
    title = "Trees Planted by Census Tract",
    subtitle = "2013–2022"
  ) +
  coord_sf(
    xlim = c(-122.53, -122.35),
    ylim = c(37.70, 37.84),
    datum = NA,
    expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

 

## Creating Final Dataset
 
```{r standardize_tract_ids}
# Make sure all census tract IDs use the same 11-digit format

building_tract_year <- building_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0"),
    year = as.numeric(year)
  )

heat_tract_year <- heat_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0"),
    year = as.numeric(year)
  )

tree_tract_year <- tree_tract_year %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0"),
    year = as.numeric(year)
  )

svi_tract <- svi_tract %>%
  mutate(
    census_tract_geoid = str_pad(census_tract_geoid, width = 11, side = "left", pad = "0")
  )

svi_zip <- svi_zip %>%
  mutate(
    zipcode = as.character(zipcode)
  )
```



```{r prepare_redlining_tract}
# Prepare redlining data at the census tract level
 
options(tigris_use_cache = TRUE)

# Get San Francisco census tract boundaries
sf_tracts_for_join <- tracts(
  state = "CA",
  county = "San Francisco",
  year = 2020,
  cb = TRUE
) %>%
  st_transform(4326) %>%
  select(
    census_tract_geoid = GEOID,
    geometry
  )

# Make sure redlining polygons are clean
sf_redlining_clean <- sf_redlining_clean %>%
  st_transform(4326) %>%
  st_make_valid()

# Turn off s2 for polygon intersection
sf_use_s2(FALSE)

# Intersect census tracts with redlining areas
redlining_tract_overlap <- st_intersection(
  sf_tracts_for_join,
  sf_redlining_clean
) %>%
  st_transform(3310) %>%
  mutate(
    overlap_area = as.numeric(st_area(geometry))
  ) %>%
  st_drop_geometry()

sf_use_s2(TRUE)

# Keep the redlining grade with the largest overlap in each tract
redlining_tract <- redlining_tract_overlap %>%
  group_by(census_tract_geoid) %>%
  slice_max(overlap_area, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    census_tract_geoid,
    dominant_redlining_grade = redlining_grade,
    dominant_redlining_category = redlining_category,
    dominant_redlining_area_id = redlining_area_id
  )

# Check result
redlining_tract %>%
  count(dominant_redlining_grade, dominant_redlining_category, sort = TRUE)

```

```{r create_tract_year_full}
# Create the full census tract-year dataset

tract_year_full <- building_tract_year %>%
  full_join(
    heat_tract_year,
    by = c("census_tract_geoid", "year")
  ) %>%
  full_join(
    tree_tract_year,
    by = c("census_tract_geoid", "year")
  ) %>%
  left_join(
    svi_tract,
    by = "census_tract_geoid"
  ) %>%
  left_join(
    redlining_tract,
    by = "census_tract_geoid"
  )

# Check final tract-year dataset
glimpse(tract_year_full)

```



Create tract-to-ZIP and tract-to-neighborhood lookup tables

```{r create_zip_neighborhood_lookup}
# Create lookup tables from census tracts to ZIP codes and neighborhoods

# For each census tract, keep the most common ZIP code
tract_zip_lookup <- final_building_data %>%
  mutate(
    census_tract_geoid = str_pad(
      as.character(round(census_tract_geoid, 0)),
      width = 11,
      side = "left",
      pad = "0"
    ),
    zipcode = as.character(zipcode)
  ) %>%
  filter(
    !is.na(census_tract_geoid),
    !is.na(zipcode)
  ) %>%
  count(census_tract_geoid, zipcode, sort = TRUE) %>%
  group_by(census_tract_geoid) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    census_tract_geoid,
    zipcode
  )


# For each census tract, keep the most common neighborhood
tract_neighborhood_lookup <- final_building_data %>%
  mutate(
    census_tract_geoid = str_pad(
      as.character(round(census_tract_geoid, 0)),
      width = 11,
      side = "left",
      pad = "0"
    ),
    neighborhood = as.character(neighborhood)
  ) %>%
  filter(
    !is.na(census_tract_geoid),
    !is.na(neighborhood),
    neighborhood != ""
  ) %>%
  count(census_tract_geoid, neighborhood, sort = TRUE) %>%
  group_by(census_tract_geoid) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    census_tract_geoid,
    neighborhood
  )

# Check data

tract_zip_lookup %>%
  head()

tract_neighborhood_lookup %>%
  head()
```


Create ZIP code-year dataset

This summarizes the tract-year dataset up to ZIP code and year.


```{r create_zip_year_full}
# Create ZIP code-year dataset

zip_year_full <- tract_year_full %>%
  left_join(
    tract_zip_lookup,
    by = "census_tract_geoid"
  ) %>%
  filter(!is.na(zipcode)) %>%
  group_by(zipcode, year) %>%
  summarise(
    total_permits = sum(total_permits, na.rm = TRUE),
    cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(
      cooling_air_updates / total_permits * 100,
      2
    ),

    median_year_built = median(median_year_built, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),

    heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    heatwave_days_2day = mean(heatwave_days_2day, na.rm = TRUE),
    heatwave_days_3day = mean(heatwave_days_3day, na.rm = TRUE),
    heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    heatwave_events_3day = mean(heatwave_events_3day, na.rm = TRUE),
    mean_apparent_temp = mean(mean_apparent_temp, na.rm = TRUE),

    trees_planted = sum(trees_planted, na.rm = TRUE),

    mean_svi_overall = mean(svi_overall, na.rm = TRUE),
    mean_svi_socioeconomic = mean(svi_socioeconomic, na.rm = TRUE),
    mean_svi_household = mean(svi_household, na.rm = TRUE),
    mean_svi_race_ethnicity = mean(svi_race_ethnicity, na.rm = TRUE),
    mean_svi_housing_transportation = mean(svi_housing_transportation, na.rm = TRUE),

    .groups = "drop"
  ) %>%
  left_join(
    svi_zip,
    by = "zipcode"
  )

# Check ZIP code-year dataset
glimpse(zip_year_full)

```

Create neighborhood-year dataset

```{r create_neighborhood_year_full}
# Create neighborhood-year dataset

neighborhood_year_full <- tract_year_full %>%
  left_join(
    tract_neighborhood_lookup,
    by = "census_tract_geoid"
  ) %>%
  filter(!is.na(neighborhood)) %>%
  group_by(neighborhood, year) %>%
  summarise(
    total_permits = sum(total_permits, na.rm = TRUE),
    cooling_air_updates = sum(cooling_air_updates, na.rm = TRUE),
    percent_cooling_air = round(
      cooling_air_updates / total_permits * 100,
      2
    ),

    median_year_built = median(median_year_built, na.rm = TRUE),
    median_building_age = median(median_building_age, na.rm = TRUE),

    heat_days_95 = mean(heat_days_95, na.rm = TRUE),
    heatwave_days_2day = mean(heatwave_days_2day, na.rm = TRUE),
    heatwave_days_3day = mean(heatwave_days_3day, na.rm = TRUE),
    heatwave_events_2day = mean(heatwave_events_2day, na.rm = TRUE),
    heatwave_events_3day = mean(heatwave_events_3day, na.rm = TRUE),
    mean_apparent_temp = mean(mean_apparent_temp, na.rm = TRUE),

    trees_planted = sum(trees_planted, na.rm = TRUE),

    mean_svi_overall = mean(svi_overall, na.rm = TRUE),
    mean_svi_socioeconomic = mean(svi_socioeconomic, na.rm = TRUE),
    mean_svi_household = mean(svi_household, na.rm = TRUE),
    mean_svi_race_ethnicity = mean(svi_race_ethnicity, na.rm = TRUE),
    mean_svi_housing_transportation = mean(svi_housing_transportation, na.rm = TRUE),

    .groups = "drop"
  )

# Check neighborhood-year dataset
glimpse(neighborhood_year_full)

```

Check final datasets

```{r check_final_geography_datasets}
# Check number of rows and years in each dataset

tract_year_full %>%
  summarise(
    geography = "Census tract",
    rows = n(),
    geographies = n_distinct(census_tract_geoid),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

zip_year_full %>%
  summarise(
    geography = "ZIP code",
    rows = n(),
    geographies = n_distinct(zipcode),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

neighborhood_year_full %>%
  summarise(
    geography = "Neighborhood",
    rows = n(),
    geographies = n_distinct(neighborhood),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )


```


```{r save_final_datasets}

# Save final  datasets
# Updated on 7/6 due to updated residential building filter (commercial building issue before)
write_csv(tract_year_full, "Data/tract_year_full.csv")
write_csv(zip_year_full, "Data/zip_year_full.csv")
write_csv(neighborhood_year_full, "Data/neighborhood_year_full.csv")
