---
title: "Historical Redlining in San Francisco"
author: "Nancy Carmona"
output: html_document
---

```{r setup, include=FALSE}
#-----setup options-----

# clear the workspace
rm(list = ls())

# knitr options, shows the code in the knitted HTML file 
knitr::opts_chunk$set(echo = TRUE)

```


```{r load_libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# load pacman, installing if needed
if (!require("pacman")) {install.packages("pacman")}

# load other packages, installing as needed
pacman::p_load(tidyverse, sf, leaflet, tigris)


```

# Introduction

This map uses historical Home Owners' Loan Corporation (HOLC) redlining data for San Francisco from the [Mapping Inequality project](https://dsl.richmond.edu/panorama/redlining/data/CA-SanFrancisco).

The HOLC assigned neighborhood grades that were used to describe perceived mortgage lending risk. These grades reflected discriminatory housing and lending practices and contributed to patterns of racial and economic inequality.

HOLC grades were assigned as:

- **A** = Best
- **B** = Still desirable
- **C** = Definitely declining
- **D** = Hazardous

## Loading Grade Data

First we load the mapping data. The file we are using is a **JSON file**, which is a common file format for storing data. JSON files can store regular data columns, like city names and HOLC grades, and they can also store spatial information, such as the coordinates needed to draw shapes on a map.

When we read this file using `read_sf()`, R recognizes it as a spatial dataset.

A spatial file should have:

1. A geometry column
2. A coordinate reference system
3. A geometry type, such as `POLYGON` or `MULTIPOLYGON`


```{r load_data, include = FALSE}
#-----load data-----

# save the file path in one object, to make it easier to edit later

redlining_file <- "~/Library/Mobile Documents/com~apple~CloudDocs/Documents/SFSU Faculty/PINC/PINC 2026/PSP-2026/Data/mappinginequality.json"

# read the spatial data 
redlining <- read_sf(redlining_file)

# check what kind of object this is
class(redlining)

# check the coordinate reference system
st_crs(redlining)

# check the geometry type
st_geometry_type(redlining)

# look at the first few rows of the data
head(redlining)

```

```{r check.data, include = FALSE}
 # select only San Francisco 
sf_redlining <- redlining %>% filter (city == "San Francisco") %>% 
  st_transform(4326)

# Note: Mapping requires longitude/latitude coordinates
# Our data are already WGS 84, but showing the transform step so you are aware


# view first rows 
head(sf_redlining)

# check the HOLC grades 
table(sf_redlining$grade, useNA = "ifany")

# remove NA
sf_redlining <- sf_redlining %>%
  filter(!is.na(grade))

# check the HOLC grades after filtering
table(sf_redlining$grade, useNA = "ifany")

```


```{r map_colors, echo = FALSE}

# Create colors for each HOLC grade
holc_colors <- c(
  "A" = alpha("green", 0.6),
  "B" = alpha("blue", 0.6),
  "C" = alpha("yellow", 0.6),
  "D" = alpha("red", 0.6)
)

# Tell leaflet how to match grades to colors
holc_palette <- colorFactor(
  palette = holc_colors,
  domain = sf_redlining$grade
)

```

```{r leaflet}

# Create the interactive map

leaflet(sf_redlining) %>%
  addTiles() %>%
  setView(
    lng = -122.4194,
    lat = 37.7749,
    zoom = 12
  ) %>%
  addPolygons(
    fillColor = ~holc_palette(grade),
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    popup = ~paste0(
      "<strong>HOLC Grade:</strong> ", grade
    )
  ) %>%
  addLegend(
    pal = holc_palette,
    values = ~grade,
    title = "HOLC Grade",
    opacity = 0.7
  )

```
