---
title: "Introduction to Statistical Analysis"
author: "Nancy Carmona & Rochelle-Jan Reyes"
date: "2025-10-24"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r set_knitr_options, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Session 6 - Introduction to Statistical Analysis in R**

In this session you will learn:

-   Utilize base R functions and packages for statistical analyses.
-   Implement different types of statistical analyses methods, including
    simple analytics and comparative tests.
-   Complete a statistical analysis in a real-world context through
    examples and demos.

## [***PPT Materials***]{.underline}

Here is the link to the Session 6 PPT. Feel free to use it as a
reference when going through this RMD file:
<https://docs.google.com/presentation/d/1dRIoI1cY5FXBbT89_Q4IsfuMQlqnlm34hY7UsrZ4NI8/edit?usp=sharing>

## [***R Functions***]{.underline}

Here is the link to the R Function Dictionary, with all of the functions
you will need for this session under the Session 5 header:
<https://docs.google.com/document/d/1z6agvbE-X70prlIdyFpU4-WkyI0IGI0WL5meEaQYz3o/edit?usp=sharing>

## A) Statistics Review

"Health statistics are numbers that summarize information related to
health." (National Library of Medicine)

-   Understand risk factors for communities

-   Track and monitor health events

-   See impacts of policy changes

-   Assess the quality and safety of health care

-   Used for evidence-informed policy making and evidence-based medicine

-   Statistics is a branch of mathematics that deals with the
    manipulation, analysis, and interpretation of data (Merriam-Webster)

-   Used in many different fields, including public health,
    epidemiology, and social sciences among other fields

Types of Statistical Studies \* Descriptive statistics \* Inferential
statistics

## [***Activity 1 - Descriptive Statistics in Base R***]{.underline}

First, as always, let's load in libraries (if any) and data.

```{r}

# load data 
ncha <- read.csv("Data/NCHA2021_Session3_5_Subset.csv")

```

Okay, now we can go through some descriptive statistics.

Remember that descriptive statistics [*describe*]{.underline} what the
data is. Inferential statistics is used to [*infer (draw)
conclusions*]{.underline} from the data.

Let's look at the dataset really quickly. How do we *view* the data?

```{r}

```

Now let's look at age. What *class* is the variable AGE?

ANSWER:

Time to use some functions we've learned in section A to describe the
data.

### Mean, Median, and Mode

First, let's look at mean (average). Which function can you use for
finding the average value in the variable AGE?

```{r}

```

*Note*: will need to include na.rm = TRUE.

Now, we can look at the median.

```{r}

```

Lastly, we can look at mode. Normally, functions of what you need are
named simply, like mean = mean() and median = median(). Mode is one of
the few that aren't. You can find the mode (most of the same number of
that variable) by using the *table()* function.

```{r}

table(ncha$AGE)

```

Is the mean, median, and mode the same number? If it's not the same
number, then the data may be skewed! We can visualize this by using the
base R plotting function, *hist()*.

```{r}


```

Which way is it skewed?

### Standard Deviation and Variance

Standard deviation (SD): how dispersed the data is in relation to the
mean

```{r}


```

Variance: expected value of the squared deviation from the mean

```{r}

```

Great job! That's it for Activity 1 in Session 6.

## B) Statistical Analyses in R

*Key Steps for your Analysis*

-   Set up analysis code

-   Exploratory Analysis

-   Data Wrangling

-   Conduct descriptive analysis

-   Conduct inferential analysis

-   Document your analysis!

## C) Demo & Activity: Applying Descriptive Statistics

Install & Load *tidyverse* package

```{r}

# load the package "tidyverse" to wrangle data

library(tidyverse)

```

Explore the NCHA dataset! What does your data structure look like?

```{r}

# View the structure of the dataset after it has been imported and saved to your working environment

glimpse(ncha)

```

Explore the NCHA dataset! What does your data structure look like?

# Tidyverse also has functions which we can use to "summarize" our data

```{r}

# we can use statistical functions and combine them together 

# summarise is a tidyverse functions 

# we can save our table in an object

ncha_summary_table <- ncha %>% 
  
  # group by and calculate summary statistics
  summarise(Min = min(MIN_MOD_PA, na.rm = TRUE), 
            Max = max(MIN_MOD_PA, na.rm = TRUE), 
            Mean = mean(MIN_MOD_PA, na.rm = TRUE),
            Median = median(MIN_MOD_PA, na.rm = TRUE), 
            missing = sum(is.na(MIN_MOD_PA)), 
            .groups = "drop") %>% 
  
  # round by mutating on a predicate function
  mutate_if(is.numeric, round, 3)

```

## D) Demo & Activity: Applying Inferential Statistics

Now, let’s say we are working on a project to examine whether the number
of minutes of moderate physical activity differs between SFSU college
students who have ever used cannabis and those who have not.

We can use data from the American College Health Association – National
College Health Assessment (ACHA-NCHA) which we used in Session 5.

I have two variables of interest:

-   "Substances ever used - Cannabis (marijuana, weed, hash, edibles,
    vaped cannabis, etc.) [nonmedical use]"

-   "Minutes of moderate physical activity in the last 7 days"

**Step 1: Data Wrangling**

Now, let’s create a data frame with our analysis variables from the main
dataset

```{r}

# Data Wrangling for Inferential Analysis

# Make subset with your variables of interest
ncha_subset <- ncha %>% select(SUB_ANY_CANNABIS, MIN_MOD_PA)


# Filter out observations that are not Yes or No & any observations that are missing minutes of moderate physical activity data
ncha_subset <- ncha_subset %>%
  filter(SUB_ANY_CANNABIS %in% c("Yes", "No")) %>%
  filter(!is.na(MIN_MOD_PA))

```

**Step 2. Descriptive Statistics**

We can look at the mean and standard deviation of each group

```{r}

# Descriptive statistics for Minutes of Moderate Physical Activity by Ever Using Cannabis 

descriptive_stats <- ncha_subset %>%
  group_by(SUB_ANY_CANNABIS) %>%
  summarize(
    Mean_Min_Mod_PA = mean(MIN_MOD_PA, na.rm = TRUE),
    SD_Min_Mod_PA = sd(MIN_MOD_PA, na.rm = TRUE),
    n = n()
  )

descriptive_stats


```

**Step 3: Conduct Statistical Test**

Now let's conduct the t-test!

```{r}

# Conduct a two sample t - test 
t.test(MIN_MOD_PA ~ SUB_ANY_CANNABIS, data = ncha_subset)

```

-   What is the p-value?

-   Are the result's statistically significant?

**Step 4. Visualize data we used in inferential test**

We can visualize our data in a boxplot using ggplot()!

```{r}

# Visualizing t-test with boxplots 

ggplot(ncha_subset, aes(x = SUB_ANY_CANNABIS, y = MIN_MOD_PA, fill =SUB_ANY_CANNABIS )) +
  geom_boxplot() +
  labs(
    title = "Moderate Physical Activity by Cannabis Status (NCHA)",
    x = "Cannabis Status",
    y = "Weekly Moderate Physical Activity (minutes)") +
         theme(legend.position = "none") 

```



# Session Information and Code

The next two chunks should be included in the appendix of every R
Markdown so that you document your session information, and code in the
document. This supports the reproducibility of your work!

```{r session.info}
#-----session information: beginning of Appendix -----------
# This promotes reproducibility by documenting the version of R and every package
# you used.
sessionInfo()
```

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, include=TRUE}
#-----appendix------------
```
