---
title: "Session3_Packages-Scripts_ActivityONLY"
author: "Rochelle-Jan Reyes"
date: "2024-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session 3 - R Sessions, Packages & Scripts, and R File Types (Importing & Exporting) --- ACTIVITIES ONLY

## Layout of the RMD Worksheets (this is a header)

This is the body of the worksheet. You can type here normally and can **bold** or *italicize* things as well as make bullet points. *This is where you can typically write your responses to questions.*

To easily create a code chunk, you will go onto a new line and type "/" and select "R Code Chunk" OR press the green square with a C and a + sign at the top right of the pane.

```{r}
#after the {r}, this is what's called a code chunk
#this is where you will enter and execute code

#for example, run this code:
2+2
```

Once you run the code by pressing Cmd+Enter (MAC), Ctrl+Enter (PC), or pressing the green play button at the top right of the code chunk, an output section will pop out at the bottom of the code chunk.

## Functions Addressed

Installing Packages

```{r}
install.packages("")

```

When you are installing a package, you need to comment it out when you are done installing OR just use the console to install it. R will run into issues when knitting if you have an install.packages() in a code chunk.

Reading in the package to the library

```{r}
library()
```

Importing in data sets / data files

```{r}
# EXCEL or TXT files
df_txt <- read.table("TXTFile_Template.txt", header = TRUE)
df_exc <- read.table("EXCELFile_Template.xlsx", header = FALSE)

# CSV files
df_csv <- read.csv("CSVFile_Template.csv", header = TRUE)
```

Accessing Infromation from the dataset

```{r}
# Viewing the data frame
View() #note that it's a capital V

# View the internal structure of the data frame
str()

# Find the number of different values in a variable
table(df$variable)
```

Exporting data sets into files

```{r}
# TXT files
write.table(mtcars, file = "mtcars.txt", sep = "\t",
            row.names = TRUE, col.names = NA)

# CSV files
write.csv(mtcars, file = "mtcars.csv")
```

## *Activity 1:* Create an RStudio Project for the Workshop

**Instructions:** 

1.  Go to *File*

2.  Select *New Project*

3.  Select *New Directory*

4.  Select *New Project*

5.  Change name to "r_onramp"

6.  Change working directory to desired computer location

7.  Press button on "Open in New Session"

8.  Create that new project!

## *Activity 2*: Importing, Viewing, and Exporting

Now it's your turn to import and export data, with a little bit of functions we learned from last session! [Remember to write your code in the pre-loaded coding blocks]{.underline}.

#### Importing the Data

First, let's import the data set. *The dataset is downloadable from Module 3's Workshop Materials on Canvas. Hint: the dataset is in a CSV file form and we need to **read** it into the Session environment.*

```{r}
#insert the code here

```

#### Working with the Data

Great! Now you have the data in your environment! The next step of this activity is to view the data set.

```{r}
# view the data set

```

Questions:

```{r}
# you can use this space to answer the questions below
```

1.  How many variables are in the data set? *Hint: you can use either View() or str() to answer this question!*

2.  What two types/classes of data are present in the data set? *Hint: use the function str() to answer the questions.*

Next, we are going to look at the values in the variable OVRL_HEALTH. How do we view the number of different values in a variable?

```{r}
# Input code here

```

Questions:

1.  How many different values does the variable OVRL_HEALTH have?

#### Creating a data frame with the current data set

We can create a data frame with data from the current data set by using the function, data.frame(). Select two or more variables from the current data set and create a new data frame.

```{r}
# Insert code below

```

#### Exporting the Data

Great! Now, let's export that data frame you made as a .txt file. *Hint: use the write.table() function.*

```{r}
# Insert code below

```

# **Great job, you finished Session 3!**

## If there's time, let's try to knit! 

-   You may also need to install Latex in order to knit with this line of code: `tinytex::install_tinytex()` in your console panel (bottom left).

-   Before you knit, make sure that all of the code that you put in the chunks works without errors.

-   Once you have installed Latex and checked for errors, you can select the Knit button at the top of this panel that has the needle and thread. Press the down arrow button and select Word or PDF.

-   R will automatically run through your code and output a knitted file. of your code, comments, and text!
